---
title: 'Clusterings Comparisons'
output:
  html_document:
    code_folding: 'hide'
---


```{r, echo=TRUE, include=FALSE}
#setwd('/afs/inf.ed.ac.uk/user/s17/s1725186/Documents/PhD-InitialExperiments/NetworkAnalysis/R_markdowns')

library(tidyverse) ; library(plotly) ; library(plotlyutils) ; 
library(RColorBrewer) ; library(viridis) ; require(gridExtra) ; library(dendextend) ; library(gplots)
library(pdfCluster)
library(WGCNA)
```
<br>

Load data

```{r load_datasets}

# Load clusterings
clusterings = read.csv('./../Data/Gandal/clusterings.csv', row.names=1)

# Gandal dataset
load('./../Data/Gandal/preprocessed_data.RData')
datExpr = datExpr[rownames(datExpr) %in% rownames(clusterings),] %>% data.frame
datGenes = datGenes %>% data.frame
DE_info = DE_info[rownames(DE_info) %in% rownames(clusterings),] %>% data.frame

# GO Neuronal annotations
GO_annotations = read.csv('./../Data/GO_annotations/genes_GO_annotations.csv')
GO_neuronal = GO_annotations %>% filter(grepl('neuron', go_term)) %>% 
              mutate('ID'=as.character(ensembl_gene_id)) %>% 
              dplyr::select(-ensembl_gene_id) %>% distinct(ID) %>%
              mutate('Neuronal'=1)


rm(GO_annotations)
```

### Compare two-class clusterings

```{r}
two_class_clusterings = clusterings %>% dplyr::select(contains('top_clusters'))
colnames(two_class_clusterings) = sub('_top_clusters', '', colnames(two_class_clusterings))
```

#### Number of unclassified genes

- Complete linkage results in the biggest ammount of unclassified gene by far

- Converting the adjacency matrix (S) to a scale free topology increases from 33 to 939 the number of unclassified genes, but using the topological overlap matrix reduces this a little (to 792)

- The Dynamic Hybrid algorithm only leaves one gene without classification while the Dynamic Tree algorithm leaves 792 out

- The euclidean distance is the distance that leaves the least amount of genes unclassified, then the biweight midcorrelation and then Pearson's correlation (but the numbers are very similar for the three)


```{r fig.width=10}
na_dist = data.frame('clustering'=colnames(two_class_clusterings), 
                     'genes'=apply(two_class_clusterings,2,function(x) sum(is.na(x))))

ggplotly(na_dist %>% ggplot(aes(reorder(clustering, -genes), genes, fill=-genes)) + geom_bar(stat='identity') + 
         ggtitle('Number of unclassified genes') + xlab('Experiment') + theme_minimal() + #scale_y_sqrt() +
         theme(legend.position='none', axis.text.x=element_text(angle=90, hjust=1)))

rm(na_dist)
```

#### Adjusted Rand Index

- No clustering seems to be significantly related to Neuronal genes

- Euclidean distance creates modules very different from the biweight and pearson correlations

```{r}
cluster_comp = two_class_clusterings %>% 
               mutate('Neuronal'=ifelse(rownames(two_class_clusterings) %in% GO_neuronal$ID,1,0))

cluster_sim = matrix(nrow = ncol(cluster_comp), ncol=ncol(cluster_comp)) %>% data.frame
for(i in 1:(ncol(cluster_comp))){
  cluster1 = cluster_comp[,i]
  for(j in (i):ncol(cluster_comp)){
    cluster2 = cluster_comp[,j]
    cluster_sim[i,j] = adj.rand.index(cluster1, cluster2)
  }
}
colnames(cluster_sim) = colnames(cluster_comp)
rownames(cluster_sim) = colnames(cluster_sim)

cluster_sim = cluster_sim %>% as.matrix %>% round(2)
heatmap.2(x = cluster_sim, Rowv = F, Colv = F, dendrogram = 'none', col=rev(brewer.pal(9,'YlOrRd'))[1:9],
          cellnote = cluster_sim, notecol = 'black', trace = 'none', key = FALSE, 
          cexRow = 1, cexCol = 1, margins = c(8,8))
 

rm(i, j, cluster1, cluster2, cluster_sim, cluster_comp)
```

#### Dendrogram

For the base dendrogram, using the final preprocessing chosen:

- Gene co-expression similarity meassure: Biweight midcorrelation

- Transformation so $S \in [0,1]$: $|S|$

- Scale free opology: power transformation with $beta=5$

- Using TOM

- Average linkage hierarchical clustering

```{r calculate_dendrogram}
allowWGCNAThreads()
S = datExpr %>% t %>% bicor %>% abs
S_sft = S^5
TOM = S_sft %>% TOMsimilarity
dissTOM = 1-TOM

rownames(dissTOM) = rownames(S_sft)
colnames(dissTOM) = colnames(S_sft)

dend = dissTOM %>% as.dist %>% hclust(method='average')

rm(S, S_sft, TOM, dissTOM)
```

- The adjusted rand index was right, the the modules between different version of the algorithm are not very similar

- The scale free topology and the topological overlap seem to increase the number of unclassified genes. This transformations have a big effect in the modules found

- The modules found in the euclidean distance are very different from the rest

- The Dynamic Hybrid algorithm classifies the lower branches very similar to the dynamic tree, but has a noisy classification of nodes close to the root

**Note:** bicor, TOM, avg_linkage and dynamic tree are actually the same clustering but have different names because they were used on different experiments.

```{r plot_dendrogram, fig.width=10, warning=FALSE}
top_module_colors = viridis(2)
dend_colors = two_class_clusterings %>% arrange(match(rownames(datExpr), labels(dend))) %>%
              mutate('S_sft'=abs(S_sft-3), 'S'=abs(S-3)) %>%
              mutate_all(function(x) ifelse(is.na(x), 'gray', top_module_colors[x])) %>%
              mutate('Neuronal'=ifelse(labels(dend) %in% GO_neuronal$ID, '#cc3399', 'gray'),
                     '----'='white','-----'='white','------'='white','-------'='white') %>%
              dplyr::select(euc, cor, bicor, `----`, S, S_sft, TOM, `-----`, comp_linkage, avg_linkage,
                            `------`, fixed_height_tree, dynamic_hybrid, dynamic_tree, `-------`, Neuronal)

par(mar=c(12,5,1,1)) # for saving the image
dend %>% as.dendrogram(hang=0) %>% set('labels', rep('', nrow(datExpr))) %>% plot(ylim=c(min(dend$height),1))
colored_bars(colors=dend_colors, y_shift=0.7) # , y_scale=0.4) # for saving the image

rm(dend_colors, top_module_colors)
```
