---
title: 'Effect of different normalisations on samples'
output:
  html_document:
    code_folding: 'hide'
---

```{r load_packages, echo=TRUE, include=FALSE}
#setwd('/afs/inf.ed.ac.uk/user/s17/s1725186/Documents/PhD-InitialExperiments/FirstYearReview/R_Markdowns')

library(tidyverse) ; library(plotly) ; require(gridExtra) ; library(ggExtra)
library(DESeq2) ; library(limma) ; library(edgeR)

create_plot = function(datExpr, title){
  plot_data = data.frame('ID'=colnames(datExpr), 'Mean'=colMeans(datExpr), 'SD'=apply(datExpr,2,sd)) %>% 
              mutate('ID'=substring(ID,2)) %>% left_join(datMeta, by=c('ID'='Dissected_Sample_ID')) %>%
              dplyr::select('ID', 'Mean', 'SD', 'Diagnosis_')
  
  p_points = plot_data %>% ggplot(aes(Mean, SD, color=Diagnosis_)) + geom_point() + theme_minimal() + 
             ggtitle(title) + theme(legend.position='bottom')
  p = ggMarginal(p_points, groupFill=T, size=3)
  
  return(p)
}
```

---

### Raw data
```{r raw_data, warning=FALSE}
load('./../Data/Gandal/filtered_raw_data.RData')

datExpr %>% create_plot('Mean expression of raw data') %>% print
```
<br>

---

### log2+1

This transformation seems to be the one that starts the heavy right tail of the ASD samples
```{r log2, warning=TRUE}
log2(datExpr+1) %>% data.frame %>% create_plot('log2 transformed data') %>% print
```

A possible explanation for this is that the right bump in the control samples was driven by outliers that were corrected by the log transformation, while the autism samples have consistently high expression in all their genes and that's why they weren't as affected by the transformation

Using the Median of expression instead of the mean because this second metric is more robust to outliers we lose the right Control bump, so it seems like that was the reason behind it after all
```{r median, warning=FALSE}
median_expr_df = data.frame('ID'=colnames(datExpr), 'Median'=apply(datExpr,2,median), 'SD'=apply(datExpr,2,sd)) %>%
               mutate('ID'=substring(ID,2)) %>% left_join(datMeta, by=c('ID'='Dissected_Sample_ID')) %>%
               dplyr::select('ID', 'Median', 'SD', 'Diagnosis_')

p_points = mean_expr_df %>% ggplot(aes(Median, SD, color=Diagnosis_)) + geom_point() + theme_minimal() + 
                            ggtitle('Median expression of raw data') + theme(legend.position='bottom')
ggMarginal(p_points, groupFill=T, size=3)

rm(median_expr_df)
```
<br>

---

## Normalisations

All normalisations seem to do very similar things with the distribution of the data
<br>

### DESeq2 package normalisations
<br>

#### vst
```{r vst, warning=FALSE}
counts = datExpr %>% as.matrix
rowRanges = GRanges(datGenes$chromosome_name,
                    IRanges(datGenes$start_position, width=datGenes$length),
                    strand=datGenes$strand,
                    feature_id=datGenes$ensembl_gene_id)
se = SummarizedExperiment(assays=SimpleList(counts=counts), rowRanges=rowRanges, colData=datMeta)
dds = DESeqDataSet(se, design = ~ Diagnosis_)
vst_norm = vst(dds) %>% assay %>% data.frame

vst_norm %>% create_plot('vst normalisation') %>% print
```
<br>

#### rlog
```{r rlog, warning=FALSE}
rlog_norm = rlog(dds) %>% assay %>% data.frame

rlog_norm %>% create_plot('rlog normalisation') %>% print

rm(counts, rowRanges, se, dds)
```
<br><br>

### limma package normalisations
<br>

#### log2 cpm
```{r cpm, warning=FALSE}
dge = DGEList(counts=datExpr, samples=datMeta, genes=datGenes)
dge = calcNormFactors(dge)

if(max(dge$samples$lib.size)/min(dge$samples$lib.size)>3){
    print('Should use voom instead of cpm because the library size ratios are too big')
}
cpm_norm = cpm(dge, log=TRUE) %>% data.frame

cpm_norm %>% create_plot('cpm normalisation') %>% print
```

#### log2 voom
```{r voom, warning=FALSE}
if(max(dge$samples$lib.size)/min(dge$samples$lib.size)<=3){
    print('Should use cpm instead of cpm because the library size ratios are not big enough')
}
voom_norm = voom(dge, log=TRUE)$E %>% data.frame

voom_norm %>% create_plot('voom normalisation') %>% print
```

