---
title: "WGCNA tutorial"
output: html_document
---

```{r, echo=TRUE, include=FALSE}
#setwd('/afs/inf.ed.ac.uk/user/s17/s1725186/Documents/PhD-InitialExperiments/NetworkAnalysis/R_markdowns')

library(tidyverse) ; library(reshape2) ; library(glue) ; library(plotly)
library(RColorBrewer) ; library(viridis) ; require(gridExtra)
library(WGCNA)
library(doParallel)

```

## Data input and cleaning
```{r}
femData = read.csv('./../Data/LiverFemale3600.csv')

datExpr0 = as.data.frame(t(femData[, -c(1:8)]))
names(datExpr0) = femData$substanceBXH
rownames(datExpr0) = names(femData)[-c(1:8)]
```

Remove outlier samples
```{r}
sampleTree = hclust(dist(datExpr0), method = "average");
sizeGrWindow(12,9)
par(cex = 0.6)
par(mar = c(0,4,2,0))
plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5, cex.axis = 1.5, cex.main = 2)
abline(h = 15, col = "red");

clust = cutreeStatic(sampleTree, cutHeight = 15, minSize = 10)
table(clust)

keepSamples = (clust==1)
datExpr = datExpr0[keepSamples, ]
nGenes = ncol(datExpr)
nSamples = nrow(datExpr)
```

Load clinical trait data

```{r}
traitData = read.csv('./../Data/ClinicalTraits.csv')

# remove columns that hold information we do not need.
allTraits = traitData[, -c(31, 16)]
allTraits = allTraits[, c(2, 11:36) ]
dim(allTraits)
names(allTraits)
# Form a data frame analogous to expression data that will hold the clinical traits.
femaleSamples = rownames(datExpr);
traitRows = match(femaleSamples, allTraits$Mice)
datTraits = allTraits[traitRows, -1]
rownames(datTraits) = allTraits[traitRows, 1]

collectGarbage()
```

## Step-by-step construction of the gene network and identification of modules

### Choosing the soft-thresholding power: analysis of network topology

```{r}
powers = c(c(1:10), seq(from = 12, to=20, by=2))
sft = pickSoftThreshold(datExpr, powerVector = powers, verbose = 5)

# Plot the results:
sizeGrWindow(9, 5)
par(mfrow = c(1,2))
cex1 = 0.8

# Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
main = paste("Scale independence"))
text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
labels=powers,cex=cex1,col="red")
abline(h=0.8, col="red") # # this line corresponds to using an R^2 cut-off of h

# Mean connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[,1], sft$fitIndices[,5],
xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
main = paste("Mean connectivity"))
text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers, cex=cex1,col="red")
```
### Co-expression similarity and adjacency
```{r}
softPower = 5
adjacency = adjacency(datExpr, power = softPower)
```

### TOM
```{r}
TOM = TOMsimilarity(adjacency)
dissTOM = 1-TOM
```

```{r}
# Call the hierarchical clustering function
geneTree = hclust(as.dist(dissTOM), method = "average");
# Plot the resulting clustering tree (dendrogram)
sizeGrWindow(12,9)
plot(geneTree, xlab="", sub="", main = "Gene clustering on TOM-based dissimilarity",
labels = FALSE, hang = 0.04)
```

```{r}
# We like large modules, so we set the minimum module size relatively high:
minModuleSize = 30
# Module identification using dynamic tree cut:
dynamicMods = cutreeDynamic(dendro = geneTree, distM = dissTOM, deepSplit = 2, 
                            pamRespectsDendro = FALSE, minClusterSize = minModuleSize)
table(dynamicMods)
```


```{r}
# Convert numeric lables into colors
dynamicColors = labels2colors(dynamicMods)
table(dynamicColors)
# Plot the dendrogram and colors underneath
sizeGrWindow(8,6)
plotDendroAndColors(geneTree, dynamicColors, "Dynamic Tree Cut",
dendroLabels = FALSE, hang = 0.03,
addGuide = TRUE, guideHang = 0.05,
main = "Gene dendrogram and module colors")
```


### Plot PCA

There doesn't seem to be a strong correlation between clusters and principal components
```{r}
library(missMDA)

nb = datExpr %>% t %>% estim_ncpPCA
res.comp = imputePCA(datExpr, ncp=nb$ncp)

pca = res.comp$completeObs %>% t %>% prcomp

plot_data = data.frame('PC1' = pca$x[,1], 'PC2' = pca$x[,2], 'cluster'=as.factor(sub(0,NA,dynamicMods)))
                       
ggplotly(plot_data %>% ggplot(aes(PC1, PC2, color=cluster)) + geom_point(alpha=0.5) + theme_minimal())
```










