---
title: 'WGCNA manually similarity measures'
output:
  html_document:
    code_folding: 'hide'
---

### Set pipeline:

- Define a gene co-expression similarity **comparing biweight midcorrelation vs correlation vs euclidean distance**

- Performing hierarchical clustering **using average linkage hclust(method='average')**

- Extracting clusters using a dynamic brach cutting approach from [Dynamic Tree Cut: in-depth description, tests and applications](https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/BranchCutting/Supplement.pdf) **using Dynamic Tree**

- Merging similar clusters using Module Eigengenes **comparing original clusters vs merged clusters vs two main clusters**

```{r, echo=TRUE, include=FALSE}
#setwd('/afs/inf.ed.ac.uk/user/s17/s1725186/Documents/PhD-InitialExperiments/NetworkAnalysis/R_markdowns')

library(tidyverse) ; library(reshape2) ; library(glue) ; library(plotly) ; library(plotlyutils)
library(RColorBrewer) ; library(viridis) ; require(gridExtra) ; library(dendextend) ; library(gplots)
library(biomaRt) ; library(DESeq2) ; library(sva) ; library(WGCNA)
library(doParallel)
library(pdfCluster)
```

Load preprocessed dataset (preprocessing code in /../FirstYearReview/data_preprocessing.Rmd)

```{r load_dataset, echo=TRUE, include=FALSE}
# Gandal dataset
load('./../../FirstYearReview/Data/Gandal/preprocessed_data.RData')
datExpr = datExpr %>% data.frame
datGenes = datGenes %>% data.frame
DE_info = DE_info %>% data.frame

# GO Neuronal annotations
GO_annotations = read.csv('./../../FirstYearReview/Data/GO_annotations/genes_GO_annotations.csv')
GO_neuronal = GO_annotations %>% filter(grepl('neuron', go_term)) %>% 
              mutate('ID'=as.character(ensembl_gene_id)) %>% 
              dplyr::select(-ensembl_gene_id) %>% distinct(ID) %>%
              mutate('Neuronal'=1)

clusterings = list()

rm(GO_annotations)
```

Keep DE genes
```{r}
datExpr = datExpr %>% filter(rownames(.) %in% rownames(DE_info)[DE_info$padj<0.05])
rownames(datExpr) = datGenes$feature_id[DE_info$padj<0.05 & !is.na(DE_info$padj)]
datGenes = datGenes %>% filter(feature_id %in% rownames(DE_info)[DE_info$padj<0.05])
DE_info = DE_info %>% filter(padj<0.05)

print(paste0('Keeping ', nrow(datExpr), ' genes and ', ncol(datExpr), ' samples'))
```
<br><br>

---

## Define a gene co-expression similarity

<br>

### Using Biweight midcorrelation

```{r}
allowWGCNAThreads()

cor_mat = datExpr %>% t %>% bicor
```


Correcting the correlation matrix from $s \in [-1,1]$ to $s \in [0,1]$ using $s_{ij}=|bw(i,j)|$

```{r}
S = abs(cor_mat)
```

### Clustering

#### Identifying gene modules

Using 1-S sa distance matrix

```{r, fig.width=10}
diss_S = 1-S
dend = diss_S %>% as.dist %>% hclust(method='average')
plot(dend, hang=0, labels=FALSE)
```

Using Dynamic Tree Cut, a dynamic branch cutting approach taken from [Dynamic Tree Cut: in-depth description, tests and applications](https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/BranchCutting/Supplement.pdf)

```{r}
modules = cutreeDynamicTree(dend, deepSplit=F, minModuleSize=20)
names(modules) = labels(dend)

clusterings[['bicor']] = modules
```

#### Merging modules with similar expression profiles

Calculate the "eigengenes" (1st principal component) of each module and merging similar modules

```{r, fig.width=10}
module_colors = c('gray', viridis(max(modules)))

MEs_output = datExpr %>% t %>% moduleEigengenes(colors=module_colors[as.vector(modules[labels(dend)])+1])
MEs = MEs_output$eigengenes
colnames(MEs) = sapply(colnames(MEs), function(x) paste0('Mod',which(module_colors == substring(x,3))) )

# Merge similar modules
bicor_dist = 1-bicor(MEs)
dend_MEs = bicor_dist %>% as.dist %>% hclust(method='average')
dend_MEs %>% as.dendrogram %>% set('labels', rep('', nrow(bicor_dist))) %>% plot(ylim=c(0, 0.6))
abline(h=0.5, col='#0099cc')
abline(h=0.2, col='#009999')

# Two main modules
tree_cut = cutree(dend_MEs, h=0.5)
top_modules = modules %>% replace(modules %in% (gsub('Mod','', names(tree_cut[tree_cut==1])) %>% as.numeric), 1) %>%
                          replace(modules %in% (gsub('Mod','', names(tree_cut[tree_cut==2])) %>% as.numeric), 2)
clusterings[['bicor_top_clusters']] = top_modules

# Closest modules
tree_cut = cutree(dend_MEs, h=0.2)
merged_modules = modules
n=0
for(i in sort(unique(tree_cut))){
  n=n+1
  merged_modules = merged_modules %>% 
                   replace(modules %in% (gsub('Mod','', names(tree_cut[tree_cut==i])) %>% as.numeric), n)
}

clusterings[['bicor_merged']] = merged_modules
```

```{r, fig.width=10}
top_module_colors = c('gray', viridis(max(top_modules)))
merged_module_colors = c('gray', viridis(max(merged_modules)))

dend_colors = data.frame('ID'=names(modules[labels(dend)]),
                         'OriginalModules' = module_colors[modules[dend$order]+1],
                         'MergedModules' = merged_module_colors[merged_modules[dend$order]+1],
                         'TopModules' = top_module_colors[top_modules[dend$order]+1])

dend %>% as.dendrogram(hang=0) %>% set('labels', rep('', nrow(diss_S))) %>% plot(ylim=c(min(dend$height),1))
colored_bars(colors=dend_colors[,-1])

rm(MEs, modules, module_colors, MEs_output, top_modules, merged_modules, tree_cut, top_module_colors, dend_colors, 
   i, diss_S)
```
<br>



### Using Pearson correlation

```{r}
cor_mat = datExpr %>% t %>% cor
```

Correcting the correlation matrix from $s \in [-1,1]$ to $s \in [0,1]$ using $s_{ij}=|bw(i,j)|$

```{r}
S = abs(cor_mat)
```

#### Clustering
```{r, fig.width=10}
diss_S = 1-S
dend = diss_S %>% as.dist %>% hclust(method='average')
plot(dend, hang=0, labels=FALSE)
```

Using Dynamic Tree Cut, a dynamic branch cutting approach taken from [Dynamic Tree Cut: in-depth description, tests and applications](https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/BranchCutting/Supplement.pdf)

```{r}
modules = cutreeDynamicTree(dend, deepSplit=F, minModuleSize=20)
names(modules) = labels(dend)

clusterings[['cor']] = modules
```

Merging modules with similar expression profiles using the "eigengenes" (1st principal component) of each module and merging similar modules

```{r, fig.width=10}
module_colors = c('gray', viridis(max(modules)))

MEs_output = datExpr %>% t %>% moduleEigengenes(colors=module_colors[as.vector(modules[labels(dend)])+1])
MEs = MEs_output$eigengenes
colnames(MEs) = sapply(colnames(MEs), function(x) paste0('Mod',which(module_colors == substring(x,3))) )

# Merge similar modules
bicor_dist = 1-bicor(MEs)
dend_MEs = bicor_dist %>% as.dist %>% hclust(method='average')
dend_MEs %>% as.dendrogram %>% set('labels', rep('', nrow(bicor_dist))) %>% plot(ylim=c(0, 0.55))
abline(h=0.5, col='#0099cc')
abline(h=0.2, col='#009999')

# Two main modules
tree_cut = cutree(dend_MEs, h=0.5)
top_modules = modules %>% replace(modules %in% (gsub('Mod','', names(tree_cut[tree_cut==1])) %>% as.numeric), 1) %>%
                          replace(modules %in% (gsub('Mod','', names(tree_cut[tree_cut==2])) %>% as.numeric), 2)
clusterings[['cor_top_clusters']] = top_modules

# Closest modules
tree_cut = cutree(dend_MEs, h=0.2)
merged_modules = modules
n=0
for(i in sort(unique(tree_cut))){
  n=n+1
  merged_modules = merged_modules %>% 
                   replace(modules %in% (gsub('Mod','', names(tree_cut[tree_cut==i])) %>% as.numeric), n)
}

clusterings[['cor_merged']] = merged_modules
```

```{r, fig.width=10}
top_module_colors = c('gray', viridis(max(top_modules)))
merged_module_colors = c('gray', viridis(length(unique(merged_modules))))

dend_colors = data.frame('ID'=names(modules[labels(dend)]),
                         'OriginalModules' = module_colors[modules[dend$order]+1],
                         'MergedModules' = merged_module_colors[merged_modules[dend$order]+1],
                         'TopModules' = top_module_colors[top_modules[dend$order]+1])

dend %>% as.dendrogram(hang=0) %>% set('labels', rep('', nrow(diss_S))) %>% plot(ylim=c(min(dend$height),1))
colored_bars(colors=dend_colors[,-1])

rm(MEs, modules, module_colors, MEs_output, top_modules, merged_modules, tree_cut, top_module_colors, dend_colors, 
   i, diss_S)
```
<br>

### Using euclidean distance

```{r}
cor_mat = datExpr %>% dist
```

Correcting the correlation matrix to $s \in [0,1]$ using $s_{ij}=\frac{dist(i,j)-min(dist)}{max(dist)-min(dist)}$

```{r}
S = (cor_mat-min(cor_mat))/(max(cor_mat)-min(cor_mat))
```

#### Clustering
```{r, fig.width=10}
dend = S %>% hclust(method='average')
plot(dend, hang=-1, labels=FALSE)
```

Using Dynamic Tree Cut, a dynamic branch cutting approach taken from [Dynamic Tree Cut: in-depth description, tests and applications](https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/BranchCutting/Supplement.pdf)

```{r}
modules = cutreeDynamicTree(dend, deepSplit=F, minModuleSize=20)
names(modules) = labels(dend)
clusterings[['euc']] = modules
```

Merging modules with similar expression profiles using the "eigengenes" (1st principal component) of each module and merging similar modules

```{r, fig.width=10}
module_colors = c('gray', viridis(max(modules)))

MEs_output = datExpr %>% t %>% moduleEigengenes(colors=module_colors[as.vector(modules[labels(dend)])+1])
MEs = MEs_output$eigengenes
colnames(MEs) = sapply(colnames(MEs), function(x) paste0('Mod',which(module_colors == substring(x,3))) )

# Merge similar modules
bicor_dist = 1-bicor(MEs)
dend_MEs = bicor_dist %>% as.dist %>% hclust(method='average')
dend_MEs %>% as.dendrogram %>% set('labels', rep('', nrow(bicor_dist))) %>% plot(ylim=c(0, 0.18))
abline(h=0.15, col='#0099cc')
abline(h=0.04, col='#009999')

# Two main modules
tree_cut = cutree(dend_MEs, h=0.15)
top_modules = modules %>% replace(modules %in% (gsub('Mod','', names(tree_cut[tree_cut==1])) %>% as.numeric), 1) %>%
                          replace(modules %in% (gsub('Mod','', names(tree_cut[tree_cut==2])) %>% as.numeric), 2)
clusterings[['euc_top_clusters']] = top_modules

# Closest modules
tree_cut = cutree(dend_MEs, h=0.04)
merged_modules = modules
n=0
for(i in sort(unique(tree_cut))){
  n=n+1
  merged_modules = merged_modules %>% 
                   replace(modules %in% (gsub('Mod','', names(tree_cut[tree_cut==i])) %>% as.numeric), n)
}

clusterings[['euc_merged']] = merged_modules
```

```{r, fig.width=10}
top_module_colors = c('gray', viridis(max(top_modules)))
merged_module_colors = c('gray', viridis(length(unique(merged_modules))))

dend_colors = data.frame('ID'=names(modules[labels(dend)]),
                         'OriginalModules' = module_colors[modules[dend$order]+1],
                         'MergedModules' = merged_module_colors[merged_modules[dend$order]+1],
                         'TopModules' = top_module_colors[top_modules[dend$order]+1])

dend %>% as.dendrogram(hang=-1) %>% set('labels', rep('', nrow(datExpr))) %>% plot(ylim=c(min(dend$height),0.6))
colored_bars(colors=dend_colors[,-1])

rm(MEs, modules, module_colors, MEs_output, top_modules, merged_modules, tree_cut, top_module_colors, dend_colors, i)
```
<br>


<br><br>

---

### Exploratory analysis of clustering

#### Adjusted Rand Index comparison

```{r adj_rand_index, fig.width = 10}
cluster_sim = data.frame(matrix(nrow = length(clusterings), ncol = length(clusterings)))
for(i in 1:(length(clusterings))){
  cluster1 = sub(0, NA, clusterings[[i]]) %>% as.factor
  for(j in (i):length(clusterings)){
    cluster2 = sub(0, NA, clusterings[[j]]) %>% as.factor
    cluster_sim[i,j] = adj.rand.index(cluster1, cluster2)
  }
}
colnames(cluster_sim) = names(clusterings)
rownames(cluster_sim) = colnames(cluster_sim)

cluster_sim = cluster_sim %>% as.matrix %>% round(2)
heatmap.2(x = cluster_sim, Rowv = F, Colv = F, dendrogram = 'none', col=rev(brewer.pal(9,'YlOrRd'))[4:9],
          cellnote = cluster_sim, notecol = 'black', trace = 'none', key = FALSE, 
          cexRow = 1, cexCol = 1, margins = c(12,12))
 

rm(i, j, cluster1, cluster2, cluster_sim)
```

#### PCA

Cluster don't follow any strong patterns, at least in the first principal components
```{r pca, warning=FALSE, fig.width=10, fig.height=8}
pca = datExpr %>% prcomp

plot_data = data.frame('ID' = rownames(datExpr), 'PC1' = pca$x[,1], 'PC2' = pca$x[,2], 
                       'PC3' = pca$x[,3], 'PC4' = pca$x[,4], 'PC5' = pca$x[,5],
                       'bicor' = sub(0,NA,clusterings[['bicor']]) %>% as.factor, 
                       'bicor_top_clusters' = sub(0,NA,clusterings[['bicor_top_clusters']]) %>% as.factor, 
                       'bicor_merged' = sub(0,NA,clusterings[['bicor_merged']]) %>% as.factor, 
                       'cor' = sub(0,NA,clusterings[['cor']]) %>% as.factor, 
                       'cor_top_clusters' = sub(0,NA,clusterings[['cor_top_clusters']]) %>% as.factor, 
                       'cor_merged' = sub(0,NA,clusterings[['cor_merged']]) %>% as.factor,
                       'euc' = sub(0,NA,clusterings[['euc']]) %>% as.factor, 
                       'euc_top_clusters' = sub(0,NA,clusterings[['euc_top_clusters']]) %>% as.factor, 
                       'euc_merged' = sub(0,NA,clusterings[['euc_merged']]) %>% as.factor)

selectable_scatter_plot(plot_data[,-1,], plot_data[,-1])

rm(pca, plot_data)
```
<br><br>
---

### Conclusions

- Using Euclidean distance creates clusters defined by their mean expression, perhaps it's not a good metric because it doesn't seem to capture any more complex behaviours in the data

- It is interesting that the Euclidean distance top clusters don't group together on the 1st PC and instead form stripes along it

- Correlation based distance metrics seem to capture a type of structure in the data not related to the first principal components

- **I think biweight midcorrelation may be the best choice because it's more robust than correlation**

<br><br>

---

#### Save clusterings

```{r save_clusterings}

clusterings_file = './../../FirstYearReview/Data/Gandal/clusterings.csv'

if(file.exists(clusterings_file)){

  df = read.csv(clusterings_file, row.names=1)
  
  if(!all(rownames(df) == rownames(datExpr))) stop('Gene ordering does not match the one in clusterings.csv!')
  
  for(clustering in names(clusterings)){
    df = df %>% mutate(!!clustering := sub(0, NA, clusterings[[clustering]]))
    rownames(df) = rownames(datExpr)
  }
  
} else {
  
  df = clusterings %>% unlist %>% matrix(nrow=length(clusterings), byrow=T) %>% t %>% data.frame %>% na_if(0)
  colnames(df) = names(clusterings)
  rownames(df) = rownames(datExpr)

}

write.csv(df, file=clusterings_file)

rm(clusterings_file, df, clustering)
```

---

#### Session info
```{r}
sessionInfo()
```
<br><br>
