---
title: 'MeSH Qualifiers Exploratory Analysis'
output:
  html_document:
    code_folding: 'hide'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=TRUE, include=FALSE}
#setwd('/afs/inf.ed.ac.uk/user/s17/s1725186/Documents/PhD-InitialExperiments/SFARI_genes/gene_network/R_markdowns')

library(tidyverse) ; library(reshape2)
library(jsonlite)
library(viridisLite) ; library(viridis) ; library(plotly) ; library(gplots) ; library(RColorBrewer)
library(Matrix)
library(igraph) ; library(visNetwork)

```

### Input

- ./../SFARI_genes_01-15-2019.csv of SFARI genes information downloaded from https://gene.sfari.org/database/gene-scoring/. Downloaded version 01-15-19.

- SFARI_scraping/genes/*gene_name*.csv file for each of the SFARI genes with the information of all the reports related to it obtained using the SFARI_scraper.py script and scrapy package.

- MeSH_terms_by_paper.json file with all the MeSH terms associated to each of the articles from the SFARI_scraper output files. Retrieved from the NCBI e-utils api using the get_articles_metadata.R script and reutils package. Downloaded on 24/05/19.

#### Get list of articles and of MeSH terms
```{r}
# Read json with MeSH terms by paper
mesh_terms_by_article_list = read_json('./../Data/MeSH_terms_by_paper.json', simplifyVector = TRUE)
```

#### Preprocess MeSH terms

- Retrieve subheaders/qualifiers by matching MeSH terms to a list of qualifiers. Differentiating MeSH terms from qualifiers with the same name by initial capital letter

```{r}
qualifiers = c('abnormalities', 'administration & dosage', 'adverse effects', 'analogs & derivatives', 'analysis', 
               'anatomy & histology', 'antagonists & inhibitors', 'biosynthesis', 'blood supply', 'cerebrospinal fluid',
               'chemical synthesis', 'chemically induced', 'classification', 'complications', 'congenital',
               'cytology', 'deficiency', 'diagnosis', 'diagnostic imaging', 'diet therapy', 'drug effects', 'drug therapy',
               'economics', 'education', 'embryology', 'enzymology', 'epidemiology', 'ethics', 'ethnology', 'etiology',
               'growth & development', 'history', 'immunology', 'injuries', 'innervation', 'instrumentation',
               'isolation & purification', 'legislation & jurisprudence', 'manpower', 'metabolism', 'methods', 'microbiology',
               'mortality', 'nursing', 'organization & administration', 'parasitology', 'pathology', 'pharmacokinetics',
               'pharmacology', 'physiology', 'physiopathology', 'poisoning', 'prevention & control', 'psychology',
               'radiation effects', 'radiotherapy', 'rehabilitation', 'secondary', 'secretion', 'standards',
               'statistics & numerical data', 'supply & distribution', 'surgery', 'therapeutic use', 'therapy', 'toxicity',
               'transmission', 'trends', 'ultrastructure', 'urine', 'utilization', 'veterinary', 'virology')

qualifiers_by_article_list = list()

for(article in names(mesh_terms_by_article_list)){
 
  mesh_terms = mesh_terms_by_article_list[[article]]
  article_qualifiers = c()
  
  if(!is.na(mesh_terms[1])){
    
    for(mesh_term in mesh_terms){

      # Look for qualifiers
      match_general = sapply(qualifiers, grepl, mesh_term)
      match_agonists = grepl('agonists', mesh_term) & !grepl('ntagonists', mesh_term)
      match_blood = grepl('blood', mesh_term) & !grepl('blood supply', mesh_term)
      match_chemistry = grepl('chemistry', mesh_term) & !grepl('immunohistochemistry', mesh_term)
      match_genetics = grepl('genetics', mesh_term) & !grepl('Xgenetics', mesh_term)
      
      # Add matches to the qualifiers list
      if(match_agonists) article_qualifiers = c(article_qualifiers, 'agonists')
      if(match_blood) article_qualifiers = c(article_qualifiers, 'blood')
      if(match_chemistry) article_qualifiers = c(article_qualifiers, 'chemistry')
      if(match_genetics) article_qualifiers = c(article_qualifiers, 'genetics')
      if(sum(match_general)>0) article_qualifiers = c(article_qualifiers, names(match_general)[match_general==TRUE])
      
    }
  
    # Update article entries
    qualifiers_by_article_list[[article]] = unique(article_qualifiers)
  }
  else qualifiers_by_article_list[[article]] = NA
  
}

rm(article, mesh_term, mesh_terms, match_general, match_agonists, match_chemistry,
   match_blood, match_genetics, article_qualifiers, mesh_terms_by_article_list)
```

```{r}
# List of all mesh terms
all_qualifiers = sort(c(qualifiers, 'agonists', 'blood', 'chemistry', 'genetics'))

print(paste0('Articles: ', length(qualifiers_by_article_list), '       Mesh terms: ', length(all_qualifiers)))

rm(qualifiers)
```

#### Create article-qualifier matrix
```{r}

all_articles = names(qualifiers_by_article_list)

article_qualifier_mat = data.frame(matrix(0, nrow=length(all_articles), ncol=sum(!is.na(all_qualifiers))))
rownames(article_qualifier_mat) = all_articles
colnames(article_qualifier_mat) = all_qualifiers[!is.na(all_qualifiers)]

for(article in all_articles){
  article_qualifiers = qualifiers_by_article_list[[article]]
  if(sum(is.na(article_qualifiers))==0){
    article_qualifier_mat[article, article_qualifiers] = 1
  }
}

rm(qualifiers_by_article_list, article, article_qualifiers)
```

An important proportion of articles (almost 10%) don't have qualifiers
```{r}
qualifiers_by_article = data.frame('id' = rownames(article_qualifier_mat), 'n_qualifiers' = rowSums(article_qualifier_mat))

bins = max(qualifiers_by_article$n_qualifiers)-min(qualifiers_by_article$n_qualifiers)+1
ggplotly(qualifiers_by_article %>% ggplot(aes(n_qualifiers)) + geom_histogram(bins=bins, alpha=0.6) + 
              ggtitle('Distribution of number of qualifiers by article') + theme_minimal())

rm(bins)
```

A few qualifiers are in no article or just a few, but there are some qualifiers present in most of the articles
```{r, warning=FALSE, fig.width=10}
articles_by_qualifier = data.frame('id' = as.character(colnames(article_qualifier_mat)),
                                   'n_articles' = colSums(article_qualifier_mat), stringsAsFactors = FALSE)

bins = max(articles_by_qualifier$n_articles)-min(articles_by_qualifier$n_articles)+1
ggplotly(articles_by_qualifier %>% ggplot(aes(n_articles)) + geom_histogram(bins=bins, alpha=0.6) + 
         scale_x_sqrt() + scale_y_sqrt() + theme_minimal() +
         ggtitle('Distribution of number of articles that contain each qualifier'))

rm(bins)
```

Many of the most popular qualifiers aren't useful
```{r, fig.width=10, fig.height=10}
top_n_counts = 50
top_qualifiers = articles_by_qualifier %>% top_n(top_n_counts, wt=n_articles)

ggplotly(top_qualifiers %>% ggplot(aes(reorder(id, n_articles), n_articles, fill=n_articles)) + 
         geom_bar(stat='identity') + scale_fill_viridis_c() + coord_flip() + 
         ggtitle(paste0('Top ', top_n_counts, ' qualifiers')) + 
         xlab('') + ylab('No. Articles') + theme_minimal() + theme(legend.position='none'))


rm(qualifiers_by_article, bins, top_n_counts)
```

#### Filters:

- Articles that have no qualifiers
```{r}
keep_articles = rowSums(article_qualifier_mat)>0
print(paste0(sum(!keep_articles), '/', nrow(article_qualifier_mat), ' articles have no qualifiers'))
```
<br>

- Qualifiers present in less than two articles

```{r}
keep_qualifiers = colSums(article_qualifier_mat)>1
print(paste0(sum(!keep_qualifiers), '/', ncol(article_qualifier_mat), ' qualifiers are mentioned in less than two articles'))

article_qualifier_mat = article_qualifier_mat[keep_articles, keep_qualifiers]
articles_by_qualifier = articles_by_qualifier %>% filter(id %in% colnames(article_qualifier_mat))

print(paste0('Remaining Articles: ', nrow(article_qualifier_mat), '       Mesh terms: ', ncol(article_qualifier_mat)))

rm(keep_articles, keep_qualifiers)
```
<br>

### Network:

#### Qualifier Network
```{r}
sparse_article_qualifier_mat = Matrix(as.matrix(article_qualifier_mat), sparse=TRUE)

# EDGES
qualifier_qualifier_mat = t(sparse_article_qualifier_mat) %*% sparse_article_qualifier_mat

edges = summary(qualifier_qualifier_mat) %>% rename('from'=i, 'to'=j, 'count'=x) %>%
        mutate(from = colnames(article_qualifier_mat)[from],
               to = colnames(article_qualifier_mat)[to]) %>%
        filter(from!=to)

# Convert count to fraction
article_qualifier_colsums = colSums(article_qualifier_mat)
names(article_qualifier_colsums) = colnames(article_qualifier_mat)
edges = edges %>% mutate('weight'=count/(article_qualifier_colsums[from]+article_qualifier_colsums[to]))

# NODES
nodes = articles_by_qualifier %>% mutate('id'=as.character(id), 'title'=as.character(id), 'value'=n_articles)

print(paste0('Nodes: ', nrow(nodes), '        Edges: ', nrow(edges)/2))
```

```{r, fig.width=10, fig.height=10}
graph = graph_from_data_frame(edges, vertices=nodes, directed=FALSE)
graph = graph %>% simplify(edge.attr.comb='first')

eigen_centrality_output = eigen_centrality(graph)
eigencentr = as.numeric(eigen_centrality_output$vector)

plot_data = data.frame('id'=graph %>% get.vertex.attribute('name'), 'eigencentrality'=eigencentr) %>%
            top_n(50, wt=eigencentrality)
ggplotly(plot_data %>% ggplot(aes(reorder(id, eigencentrality), eigencentrality, fill=eigencentrality)) + 
         geom_bar(stat='identity') + scale_fill_viridis_c() + coord_flip() + 
         ggtitle(paste0('Qualifiers with the highest Network centrality')) + 
         xlab('') + ylab('Eigencentrality') + theme_minimal() + theme(legend.position='none'))

rm(plot_data)
```

#### Clustering
```{r, fig.width=10}

comms_louvain = cluster_louvain(graph)
modules = comms_louvain %>% membership

color_pal = viridis_pal()(max(modules)) %>% substr(1,7)
graph = graph %>% set_vertex_attr('eigencentrality', value=eigencentr) %>%
                  set_vertex_attr('module', value=as.numeric(modules)) %>% 
                  set_vertex_attr('color_modules', value=color_pal[as.numeric(modules)]) %>%
                  set_vertex_attr('color', value=color_pal[as.numeric(modules)])

comm_sizes = sizes(comms_louvain) %>% data.frame %>% rename(Module=Community.sizes, size=Freq)
ggplotly(comm_sizes %>% ggplot(aes(Module, size, fill=size)) + geom_bar(stat='identity') + 
         ggtitle('Number of Qualifiers in each Module') + theme_minimal() + theme(legend.position = 'none'))

visIgraph(graph) %>% visOptions(selectedBy='module', highlightNearest=TRUE) %>% 
                     visIgraphLayout(randomSeed=123)


rm(color_pal, eigen_centrality_output, eigencentr, comm_sizes, edges, nodes)
```

Most important qualifiers for each module
```{r, fig.width=10}

module_info = tibble('module'=character(), 'top_term'=character(), size=integer())

for(i in names(sizes(comms_louvain))){
  
  module_vertices = names(modules)[modules==i]
  
  # Creat subgraph
  module_graph = induced_subgraph(graph, module_vertices)
  
  # Calculate eigencentrality
  eigen_centrality_output = eigen_centrality(module_graph)
  eigencentr = as.numeric(eigen_centrality_output$vector)
  names(eigencentr) = module_vertices
  
  module_info = module_info %>% add_row('module'=i, 'top_term'=names(eigencentr)[eigencentr==max(eigencentr)][1], 
                                        'size'=length(module_vertices))
    
  # Plot eigencentrality of top 10 MeSH terms
  plot_data = data.frame('qualifier'=names(eigencentr), 'eigencentrality'=eigencentr) %>%
              top_n(10, wt=eigencentrality)
  print(plot_data %>% ggplot(aes(reorder(qualifier, eigencentrality), eigencentrality, fill=eigencentrality)) + 
       geom_bar(stat='identity') + scale_fill_viridis_c() + coord_flip() + 
       ggtitle(paste0('Top terms for Module ', i)) + 
       xlab('Qualifiers') + ylab('Eigencentrality') + theme_minimal() + theme(legend.position='none'))

}

rm(i, module_vertices, module_graph, eigen_centrality_output, eigencentr, plot_data, comms_louvain)
```

#### Module Network

- Collapsing edges: sum all original weights and divide by the product of the number of elements in both (the number of possible interactions they could have)

- The Network plot is not that informative

```{r}

module_graph = contract(graph, graph %>% get.vertex.attribute('module'), vertex.attr.comb='first') %>%
               simplify(edge.attr.comb='sum') %>% set_vertex_attr('name', value=module_info$module) %>%
                                                  set_vertex_attr('title', value=module_info$top_term) %>%
                                                  set_vertex_attr('module_size', value=module_info$size) %>%
                                                  set_vertex_attr('value', value=module_info$size)

nodes_module_graph  = module_graph %>% as_data_frame(what='vertices') %>% mutate('id'=name)
edges_module_graph = module_graph %>% as_data_frame() %>% 
                     mutate('corrected_weight' = weight/(nodes_module_graph$module_size[as.numeric(from)]*
                                                 nodes_module_graph$module_size[as.numeric(to)])*100)

module_graph = module_graph %>% set_edge_attr('weight', value=edges_module_graph$corrected_weight)

visIgraph(module_graph) %>% visIgraphLayout(randomSeed=123) %>% visOptions(highlightNearest=TRUE)

rm(module_info, nodes_module_graph, edges_module_graph)
```

```{r, fig.width=10}
adj_mat = module_graph %>% as_adjacency_matrix(attr='weight') %>% as.matrix
adj_mat = adj_mat

heatmap.2(adj_mat, cellnote=round(adj_mat,2), dendrogram='none', notecol='gray', scale='none', trace='none',
          key=FALSE, xlab='Modules', ylab='Modules', col=colorRampPalette(brewer.pal(8, 'RdPu'))(25))

rm(adj_mat)
```

Save file with the modules each paper belongs to through its qualifiers
```{r}

qualifier_module = graph %>% as_data_frame(what='vertices') %>% dplyr::select(name, module)

qualifier_module_mat = matrix(0, nrow(qualifier_module), max(qualifier_module$module))
rownames(qualifier_module_mat) = qualifier_module$name
colnames(qualifier_module_mat) = sort(unique(qualifier_module$module))

for(i in seq(1, nrow(qualifier_module))){
  qualifier_module_mat[i,qualifier_module[i,2]] = 1
}

if(!all(colnames(article_qualifier_mat) == rownames(qualifier_module_mat))){
  print('Row and column order dont match!')
}

article_module_mat = as.matrix(article_qualifier_mat) %*% qualifier_module_mat
rownames(article_module_mat) = rownames(article_qualifier_mat)

#write.csv(article_module_mat, file='./../Data/article_module_matrix_qualifiers.csv')

print('Distribution of number of modules per article')
table(apply(article_module_mat, 1, function(x) sum(x>0)))
```

```{r, fig.height=8}

melt_article_module = melt(article_module_mat)
colnames(melt_article_module) = c('articleID', 'module', 'value')
melt_article_module$module = as.factor(melt_article_module$module)

ggplotly(melt_article_module %>% ggplot(aes(value, fill=module, color=module)) +
         geom_histogram(position='identity', bins=max(melt_article_module$value), alpha=0.5) + 
         facet_grid(module~., scales='free_y') + ggtitle('Module content distribution per article') + 
         xlab('n_articles') + ylab('Module content') + theme_minimal() + theme(legend.position='none'))

rm(qualifier_module, qualifier_module_mat, i)
```