---
title: 'Gene-Year of Publication Analysis'
output:
  html_document:
    code_folding: 'hide'
---

```{r, echo=TRUE, include=FALSE}
#setwd('/afs/inf.ed.ac.uk/user/s17/s1725186/Documents/PhD-InitialExperiments/SFARI_genes/gene_network/R_markdowns')

library(tidyverse) ; library(reshape2)
library(jsonlite)
library(viridisLite) ; library(viridis) ; library(plotly) ; library(gplots) ; library(RColorBrewer)
library(Matrix)
library(igraph) ; library(visNetwork)

SFARI_colour_hue = function() pal = c('#FF7631','#FFB100','#E8E328','#8CC83F','#62CCA6','#59B9C9','#b3b3b3')
```
<br><br>

### Input

- ./../SFARI_genes_01-15-2019.csv of SFARI genes information downloaded from https://gene.sfari.org/database/gene-scoring/. Downloaded version 01-15-19.

- SFARI_scraping/genes/*gene_name*.csv file for each of the SFARI genes with the information of all the reports related to it obtained using the SFARI_scraper.py script and scrapy package.

- MeSH_terms_by_paper.json file with all the MeSH terms associated to each of the articles from the SFARI_scraper output files. Retrieved from the NCBI e-utils api using the get_articles_metadata.R script and reutils package. Downloaded on 24/05/19.

- year_by_paper.csv file with the year of publication of every paper

<br><br><br>

### Preprocessing

#### Create gene-article data frame
```{r}

### GET LIST OF GENES
list_gene_files = list.files('./../SFARI_scraping/genes/')


# GET LIST OF PUBMED IDS
get_pubmed_IDs = function(){
  all_pubmed_ids = c()
  for(gene_file in list_gene_files){
    file = read.delim(paste0('./../SFARI_scraping/genes/', gene_file))
    pubmed_ids = unique(file$pubmed.ID)
    all_pubmed_ids = unique(c(all_pubmed_ids, pubmed_ids)) 
  }
  all_pubmed_ids = as.character(all_pubmed_ids)
  return(all_pubmed_ids)
}

all_pubmed_ids = get_pubmed_IDs()


# CREATE GENE-ARTICLE DATA FRAME
SFARI_genes_info = read.csv('./../../Data/SFARI_genes_01-15-2019.csv') %>% mutate('ID'=gene.symbol)

create_gene_pmID_df = function(){
  
  gene_pmID_list = list()
  
  # Fillout list
  for(gene_file in list_gene_files){
    file = read.delim(paste0('./../SFARI_scraping/genes/', gene_file))
    pubmed_ids = as.character(unique(file$pubmed.ID))
    gene_pmID_list[[gsub('.csv', '', gene_file)]] = pubmed_ids
  }
  
  gene_pmID_df = gene_pmID_list %>% stack %>% rename('gene'=ind, 'pubmedID'=values) %>% select(gene, pubmedID)
    
  return(gene_pmID_df)
}

gene_pmID_df = create_gene_pmID_df()

print(paste0('Genes: ', length(unique(gene_pmID_df$gene)), '     Articles: ', length(unique(gene_pmID_df$pubmedID))))

rm(list_gene_files)
```

#### Create article-year data frame
```{r, warning=FALSE, fig.width=10}
year_paper_df = read.csv('./../Data/year_by_paper.csv')

plot_data = year_paper_df %>% dplyr::select(year) %>% group_by(year) %>% count

print(paste0('Articles: ', nrow(year_paper_df), '     Years: ', nrow(plot_data)))

ggplotly(plot_data %>% ggplot(aes(year, n, fill=n)) + geom_bar(stat='identity') + 
         ggtitle('Publication Years') + theme_minimal() + coord_flip())

rm(plot_data)
```

#### Create gene-year data frame

Even though there are more 2018 papers than 2017, they are referenced left often. Perhaps the SFARI gene database should be updated?

Most referenced articles:

1. https://www.ncbi.nlm.nih.gov/pubmed/25363768: The contribution of de novo coding mutations to autism spectrum disorder

2. https://www.ncbi.nlm.nih.gov/pubmed/25363760: Synaptic, transcriptional and chromatin genes disrupted in autism

3. https://www.ncbi.nlm.nih.gov/pubmed/25533962: Large-scale discovery of novel genetic causes of developmental disorders

```{r, warning=FALSE, fig.width=10}
gene_year_df = gene_pmID_df %>% mutate('pubmedID'=as.integer(pubmedID)) %>% 
               left_join(year_paper_df, by='pubmedID') %>%
               left_join(SFARI_genes_info, by=c('gene'='ID')) %>%
               dplyr::select(gene, pubmedID, year, gene.score, syndromic) %>%
               mutate('gene.score'=as.factor(gene.score))

plot_data = gene_year_df %>% group_by(year, pubmedID) %>% count

ggplotly(plot_data %>% ggplot(aes(year, n, fill=n, id=pubmedID)) + geom_bar(stat='identity') + coord_flip() + 
         ggtitle('Mentions of each paper in the SFARI gene database by year') + theme_minimal())

rm(plot_data)
```

All gene groups seem to follow similar patterns
```{r, warning=FALSE, fig.width=10}
bins = max(gene_year_df$year, na.rm = TRUE)-min(gene_year_df$year, na.rm = TRUE)+1
ggplotly(gene_year_df %>% ggplot(aes(year, fill=gene.score)) + geom_histogram(position='identity', bins=bins) + 
         facet_grid(gene.score~.) + scale_fill_manual(values=SFARI_colour_hue(), na.value='#b3b3b3') +
         ggtitle('Publication Years by gene preserving scale') + theme_minimal() + theme(legend.position='none'))
```

The research of genes with lower SFARI scores seems to have decreased in the last years, specially score 6, and the researched has increased for the genes with the highest scores. This could be making the attention bias towards higher scored genes bigger.
```{r, warning=FALSE, fig.width=10}
ggplotly(gene_year_df %>% ggplot(aes(year, fill=gene.score)) + geom_histogram(position='identity', bins=bins) + 
         facet_grid(gene.score~., scales='free_y') + scale_fill_manual(values=SFARI_colour_hue(), na.value='#b3b3b3') +
         ggtitle('Publication Years by gene preserving scale adjusting scales') + theme_minimal() + theme(legend.position='none'))

rm(bins)
```

### Network Analysis

#### Create gene-year matrix
```{r}
gene_year_df_summ = gene_year_df %>% group_by(gene, year) %>% count %>% ungroup %>% ungroup %>% filter(complete.cases(.))

# Create matrix
gene_year_mat = matrix(0, nrow=length(unique(gene_year_df_summ$gene)), ncol=length(unique(gene_year_df_summ$year)))
rownames(gene_year_mat) = unique(gene_year_df_summ$gene)
colnames(gene_year_mat) = sort(unique(gene_year_df_summ$year))

# Fill matrix
gene_year_df_summ = gene_year_df_summ %>% as.matrix
for(i in 1:nrow(gene_year_df_summ)) gene_year_mat[gene_year_df_summ[i,1], gene_year_df_summ[i,2]] = gene_year_df_summ[i,3]

gene_year_mat = Matrix(gene_year_mat, sparse = TRUE)

rm(gene_year_df_summ, i, gene_year_df)
```

#### Create Network

On average, each gene is connected to a third of the genes by the year of publication of their articles
```{r}
# EDGES
gene_gene_mat = gene_year_mat %*% t(gene_year_mat)
gene_names = rownames(gene_year_mat)

edges = summary(gene_gene_mat) %>% rename('from'=i, 'to'=j, 'count'=x) %>% 
        mutate(from = gene_names[from], to = gene_names[to]) %>%
        filter(from!=to)

# Convert count to fraction
gene_year_mat_rowsums = rowSums(gene_year_mat)
names(gene_year_mat_rowsums) = rownames(gene_gene_mat)
edges = edges %>% mutate('weight'=count/(gene_year_mat_rowsums[from]+gene_year_mat_rowsums[to]))

# NODES
nodes = SFARI_genes_info %>% dplyr::select(gene.symbol, gene.score, syndromic, number.of.reports) %>% 
        mutate(id = gene.symbol, title=gene.symbol, value=number.of.reports,
               shape=ifelse(syndromic==0, 'circle','square'),
               color_score=case_when(gene.score==1~'#FF7631', gene.score==2~'#FFB100',
                                     gene.score==3~'#E8E328', gene.score==4~'#8CC83F',
                                     gene.score==5~'#62CCA6', gene.score==6~'#59B9C9',
                                     is.na(gene.score) ~ '#b3b3b3')) %>%
        mutate('color'=color_score) %>% filter(!duplicated(id)) %>% 
        filter(gene.symbol %in% unique(c(edges$from, edges$to)))

# Details:
print(paste0('Nodes: ', nrow(nodes),'        Edges: ', nrow(edges)/2))
```

#### Calculate eigencentrality by Gene Score

The higher the SFARI score, the higher the eigencentrality of the genes in the Network. Same with syndromic tag
```{r, fig.width=10}
graph = graph_from_data_frame(edges, vertices=nodes, directed=FALSE)
graph = graph %>% simplify(edge.attr.comb='first')

eigen_centrality_output = eigen_centrality(graph)
eigencentr = as.numeric(eigen_centrality_output$vector)
graph = graph %>% set_vertex_attr('eigencentrality', value=eigencentr)

plot_data = data.frame('gene.score'= as.factor(vertex_attr(graph, 'gene.score')),
                       'Syndromic'= as.logical(vertex_attr(graph, 'syndromic')),
                       'Eigencentrality'= vertex_attr(graph, 'eigencentrality'))

correlation_scr = cor(nodes$gene.score[!is.na(nodes$gene.score)], eigencentr[!is.na(nodes$gene.score)])
correlation_syn = cor(as.numeric(plot_data$Syndromic), plot_data$Eigencentrality)

p1 = ggplotly(plot_data %>% ggplot(aes(gene.score, Eigencentrality, fill=gene.score)) + geom_boxplot() + 
         scale_fill_manual(values=SFARI_colour_hue()) + theme_minimal() + theme(legend.position='none'))

p2 = ggplotly(plot_data %>% ggplot(aes(Syndromic, Eigencentrality, fill=Syndromic)) + geom_boxplot() + 
     theme_minimal() + theme(legend.position='none') + 
     ggtitle(paste0('Correlation = ', round(correlation_scr, 2), ' (left) and ', round(correlation_syn,2), ' (right)')))

subplot(p1, p2, nrows=1, widths=c(0.7, 0.3))

rm(correlation_scr, correlation_syn, p1, p2)
```

When separating the SFARI scores by syndromic tag, the relation between eigencentrality and score persists
```{r, fig.width=10}
corr_non_syn = cor(nodes$gene.score[!is.na(nodes$gene.score) & nodes$syndromic==0], 
                   eigencentr[!is.na(nodes$gene.score) & nodes$syndromic==0])
corr_syn = cor(nodes$gene.score[!is.na(nodes$gene.score) & nodes$syndromic==1], 
               eigencentr[!is.na(nodes$gene.score) & nodes$syndromic==1])
ggplotly(plot_data %>% ggplot(aes(gene.score, Eigencentrality, fill=gene.score)) + geom_boxplot() + 
         scale_fill_manual(values=SFARI_colour_hue()) + theme_minimal() + theme(legend.position='none') +
         facet_wrap( ~ Syndromic, ncol=2) + ggtitle(paste0('Correlation = ', round(corr_non_syn, 4),
                                                           ' (left) and ', round(corr_syn, 4), ' (right)')))

rm(plot_data, corr_non_syn, corr_syn, nodes, edges)
```

Most of the central nodes belong to SFARI score 1
```{r}
plot_data = data.frame('id'=graph %>% get.vertex.attribute('name'), 'eigencentrality'=eigencentr) %>%
            left_join(SFARI_genes_info, by=c('id'='ID')) %>% top_n(25, wt=eigencentrality)

ggplotly(plot_data %>% ggplot(aes(reorder(id, eigencentrality), eigencentrality, fill=as.factor(gene.score))) + 
         geom_bar(stat='identity') + coord_flip() + scale_fill_manual(values=SFARI_colour_hue(), na.value='#b3b3b3') +
         ggtitle(paste0('Top 25 genes with the highest Network centrality')) + xlab('') + ylab('Eigencentrality') + 
         theme_minimal() + theme(legend.position='none'))

rm(plot_data)
```

<!-- ```{r, fig.width=10, fig.height=6} -->
<!-- visIgraph(graph) %>% visOptions(selectedBy='gene.score', highlightNearest=TRUE) %>%  -->
<!--                               visIgraphLayout(randomSeed=123) -->
<!-- ``` -->
**Note:** Network is a bit heavy and not very informative

#### Community detection:

```{r, warning=FALSE, error=FALSE}

comms_louvain = cluster_louvain(graph)
modules = comms_louvain %>% membership

color_pal = viridis_pal()(max(modules)) %>% substr(1,7)
graph = graph %>% set_vertex_attr('module', value=as.numeric(modules)) %>% 
                  set_vertex_attr('color_modules', value=color_pal[as.numeric(modules)]) %>%
                  set_vertex_attr('color', value=color_pal[as.numeric(modules)])

comm_sizes = sizes(comms_louvain) %>% data.frame %>% rename(Module=Community.sizes, size=Freq)
ggplotly(comm_sizes %>% ggplot(aes(Module, size, fill=size)) + geom_bar(stat='identity') + 
         ggtitle('Number of genes in each module') + theme_minimal() + theme(legend.position = 'none'))

hm_data = table(vertex_attr(graph, 'gene.score'), vertex_attr(graph, 'module'), useNA='ifany') %>% 
          as.data.frame.matrix %>% as.matrix
heatmap.2(hm_data, cellnote=hm_data, notecol='white', Rowv=F, Colv=F, scale='row', trace='none', 
          key=FALSE, xlab='Module', ylab='SFARI score', col=colorRampPalette(brewer.pal(8, 'RdPu'))(25))

hm_data = table(vertex_attr(graph, 'syndromic'), vertex_attr(graph, 'module'), useNA='ifany') %>% 
          as.data.frame.matrix %>% as.matrix
heatmap.2(hm_data, cellnote=hm_data, notecol='white', Rowv=F, Colv=F, scale='row', trace='none', 
          key=FALSE, xlab='Module', ylab='Syndromic', col=colorRampPalette(brewer.pal(8, 'RdPu'))(25))
```

Separating by syndromic tag the SFARI scores
```{r, warning=FALSE}
hm_data = data.frame('syndromic_gene.score'=paste(vertex_attr(graph, 'syndromic'), '_',vertex_attr(graph, 'gene.score')),
                     'module'=vertex_attr(graph, 'module'))
hm_data = table(hm_data$syndromic_gene.score, hm_data$module) %>% as.data.frame.matrix %>% as.matrix
heatmap.2(hm_data, cellnote=hm_data, notecol='white', scale='row', trace='none', 
          key=FALSE, xlab='Module', ylab='Syndromic tag + SFARI Score', col=colorRampPalette(brewer.pal(8, 'RdPu'))(25))

rm(color_pal, hm_data)
```

#### Module characterisation

Most important genes for each module coloured by SFARI score
```{r, warning=FALSE, fig.height=3}

module_info = tibble('module'=character(), 'top_term'=character(), size=integer())

for(i in names(sizes(comms_louvain))){
  
  module_vertices = names(modules)[modules==i]
  
  # Creat subgraph
  module_graph = induced_subgraph(graph, module_vertices)
  
  # Calculate eigencentrality
  eigen_centrality_output = eigen_centrality(module_graph)
  eigencentr = as.numeric(eigen_centrality_output$vector)
  names(eigencentr) = module_vertices
  
  module_info = module_info %>% add_row('module'=i, 'top_term'=names(eigencentr)[eigencentr==max(eigencentr)][1], 
                                        'size'=length(module_vertices))
    
  # Plot eigencentrality of top 10 genes
  plot_data = data.frame('gene'=names(eigencentr), 'eigencentrality'=eigencentr) %>%
              top_n(10, wt=eigencentrality) %>% left_join(SFARI_genes_info, by=c('gene'='ID'))
  print(plot_data %>% ggplot(aes(reorder(gene, eigencentrality), eigencentrality, fill=as.factor(gene.score))) + 
       geom_bar(stat='identity') + scale_fill_manual(values=SFARI_colour_hue(), na.value='#b3b3b3') + coord_flip() + 
       ggtitle(paste0('Top genes for Module ', i)) + xlab('Genes') + ylab('Eigencentrality') + 
       theme_minimal() + theme(legend.position='none'))

}

rm(i, module_vertices, module_graph, eigen_centrality_output, eigencentr, plot_data)
```

Most important years for each module
```{r, fig.height=8, warning=FALSE}
for(i in names(sizes(comms_louvain))){
  
  module_genes = names(modules)[modules==i]
  
  # Filter genes belonging to module and column with non-zero values
  module_gene_year_mat = gene_year_mat[rownames(gene_year_mat) %in% module_genes,]
  module_gene_year_mat = module_gene_year_mat[,colSums(module_gene_year_mat)>0]
  
  # Do PCA and extract 1st PC
  pca_module = prcomp(module_gene_year_mat, scale.=TRUE)
  module_eigengene = pca_module$rotation[,'PC1']
  names(module_eigengene) = colnames(module_gene_year_mat)
  
  # Plot module's year relevance
  plot_data = data.frame('year'=names(module_eigengene), 'eigencentrality'=module_eigengene) %>%
              right_join(data.frame('year'=factor(sort(unique(year_paper_df$year), decreasing=T))), by='year') %>%
              mutate('eigencentrality'=ifelse(is.na(eigencentrality), 0, eigencentrality))
  
  print(plot_data %>% ggplot(aes(factor(year), eigencentrality, fill=abs(eigencentrality))) + 
       geom_bar(position='identity', stat='identity') + scale_fill_viridis_c() + coord_flip() + 
       ggtitle(paste0('Top Years for Module ', i)) + 
       xlab('Years') + ylab('Eigencentrality') + theme_minimal() + theme(legend.position='none'))
  
}

#rm(i, module_genes, module_gene_year_mat, pca_module, module_eigengene, plot_data)
```

