---
title: 'SFARI gene network for ASD papers'
output:
  html_document:
    code_folding: 'hide'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=TRUE, include=FALSE}
#setwd('/afs/inf.ed.ac.uk/user/s17/s1725186/Documents/PhD-InitialExperiments/SFARI_genes/gene_network/R_markdowns')

library(tidyverse)
library(plotly)
library(Matrix)
library(visNetwork)
library(igraph)

SFARI_colour_hue = function() pal = c('#FF7631','#FFB100','#E8E328','#8CC83F','#62CCA6','#59B9C9','#b3b3b3')
```
<br><br>

### Input:

- .csv of SFARI genes information downloaded from https://gene.sfari.org/database/gene-scoring/. Downloaded version 01-15-19.

- .csv file for each of the SFARI genes with the information of all the reports related to it obtained from the SFARI website https://gene.sfari.org/database/human-gene/gene_name using scrapy. Date of web scraping: 20/05/19, script used: ./SFARI_scraping/SFARI_scraping/spiders/SFARI_scraper.py

- For each gene, filtered reports to keep only the ones related to autism.

<br><br><br>

#### Get list of genes and list of pubmed ids
```{r}

# Select if results should be filtered to ASD related or not
filter_ASD = TRUE

### Get list of genes and pubmed IDs

# Genes
list_gene_files = list.files('./../SFARI_scraping/genes/')

# Pubmed IDs
get_pubmed_IDs = function(autism=FALSE){
  all_pubmed_ids = c()
  n = 0
  
  for(gene_file in list_gene_files){
    file = read.delim(paste0('./../SFARI_scraping/genes/', gene_file))
    
    if(autism) file = file %>% filter(Autism.Report == 'Yes')
    
    if(nrow(file)>0){
      pubmed_ids = unique(file$pubmed.ID)
      all_pubmed_ids = unique(c(all_pubmed_ids, pubmed_ids)) 
    }
    
    n = c(n, length(all_pubmed_ids))
  }
  
  all_pubmed_ids = as.character(all_pubmed_ids)
  
  return(list('n'=n, 'all_pubmed_ids'=all_pubmed_ids))
}

get_pubmed_IDs_output = get_pubmed_IDs(filter_ASD)
n = get_pubmed_IDs_output$n
all_pubmed_ids = get_pubmed_IDs_output$all_pubmed_ids
```

```{r}

# See change in number of pubmed articles by each new gene (Seems like the number of new articles were beginning to decrease in the end)
pubmed_counts = data.frame('genes' = seq(1,length(n)), 'articles'=n)

ggplotly(pubmed_counts %>% ggplot(aes(genes, articles,color='#434343')) + geom_line() + coord_fixed() +
  geom_smooth(method='lm', color='#666666', size=0.5) + theme_minimal() + theme(legend.position='none') +
  ggtitle('Increase in number of pubmed articles by each new gene'))

rm(n, get_pubmed_IDs_output)
```

#### Create gene-pubmed ID (sparse) matrix
```{r, fig.width=10}

# Create gene - pubmedID sparse matrix
SFARI_genes_info = read.csv('./../../Data/SFARI_genes_01-15-2019.csv') %>% mutate('ID'=gene.symbol)

create_gene_pmID_mat = function(autism=FALSE){
  
  # Create empty dataframe
  gene_pmID_mat = data.frame(matrix(0, nrow=length(list_gene_files), ncol=length(all_pubmed_ids)))
  rownames(gene_pmID_mat) = sapply(list_gene_files, function(x) gsub('.csv', '', x))
  colnames(gene_pmID_mat) = all_pubmed_ids
  
  # Fillout dataframe
  for(gene_file in list_gene_files){
    file = read.delim(paste0('./../SFARI_scraping/genes/', gene_file))
    
    if(autism) file = file %>% filter(Autism.Report == 'Yes')
    
    if(nrow(file)>0){
      pubmed_ids = as.character(unique(file$pubmed.ID))
      gene_pmID_mat[gsub('.csv', '', gene_file), pubmed_ids] = 1 
    }
  }
  
  gene_names = rownames(gene_pmID_mat)
  sparse_gene_pmID_mat = Matrix(as.matrix(gene_pmID_mat), sparse=TRUE)
    
  return(list('gene_names'=gene_names, 'sparse_gene_pmID_mat'=sparse_gene_pmID_mat))
}

gene_pmID_mat_output = create_gene_pmID_mat(filter_ASD)
gene_names = gene_pmID_mat_output$gene_names
gene_pmID_mat = gene_pmID_mat_output$sparse_gene_pmID_mat

plot_data = data.frame('ID' = rownames(gene_pmID_mat), 'n_papers'=rowSums(gene_pmID_mat)) %>% 
            left_join(SFARI_genes_info, by='ID') %>% mutate(gene.score=as.factor(gene.score))

bins = max(plot_data$n_papers)-min(plot_data$n_papers)+1
p1 = ggplotly(plot_data %>% ggplot(aes(n_papers)) + geom_histogram(bins=bins, alpha=0.6) +
         theme_minimal() + ggtitle('Distribution of number of papers per gene'))

p2 = ggplotly(plot_data %>% ggplot(aes(gene.score, n_papers, fill=gene.score)) + geom_boxplot() + 
              scale_fill_manual(values=SFARI_colour_hue()) + theme_minimal() + 
              ggtitle('Number of papers per gene (left) and by SFARI score (right)'))

subplot(p1, p2, nrows=1)
```

```{r}
# Remove genes with no articles related to them
if(sum(rowSums(gene_pmID_mat)==0)>0){
  print(paste0('Removing ', sum(rowSums(gene_pmID_mat)==0), ' rows with zero counts from gene-pubmedID count matrix'))
  gene_pmID_mat = gene_pmID_mat[rowSums(gene_pmID_mat)>0,] 
}
```

#### Create node and edge list for Network
```{r}
# Nodes
nodes = SFARI_genes_info %>% dplyr::select(gene.symbol, gene.score, syndromic, number.of.reports) %>% 
        mutate(id = gene.symbol, title=gene.symbol, value=number.of.reports,
               shape=ifelse(syndromic==0, 'circle','square'),
               color=case_when(gene.score==1 ~ '#FF7631', gene.score==2 ~ '#FFB100',
                               gene.score==3 ~ '#E8E328', gene.score==4 ~ '#8CC83F',
                               gene.score==5 ~ '#62CCA6', gene.score==6 ~ '#59B9C9',
                               is.na(gene.score) ~ '#b3b3b3')) %>%
        filter(!duplicated(id))


# Edges
edges = tibble('from'=character(), 'to'=character(), 'weight'=integer())

gene_gene_mat = gene_pmID_mat %*% t(gene_pmID_mat)
edges = summary(gene_gene_mat) %>% rename('from'=i, 'to'=j, 'weight'=x) %>% 
        mutate(from = gene_names[from], to = gene_names[to]) %>%
        filter(from!=to)

nodes = nodes %>% filter(gene.symbol %in% unique(c(edges$from, edges$to)))

# Details:
print(paste0('Nodes: ', nrow(nodes),'        Edges: ', nrow(edges)))

print(paste0('Lost ', length(unique(SFARI_genes_info$gene.symbol[!SFARI_genes_info$gene.symbol %in% nodes$id])),
             ' genes because they didn\'t share any publication with other genes'))
```

#### Scores of the genes lost: 

Lost genes seem to come from all scores proportionally
```{r}
table(SFARI_genes_info$gene.score[!SFARI_genes_info$gene.symbol %in% nodes$gene.symbol], useNA='ifany')
```

Count of original genes by score:
```{r}
table(SFARI_genes_info$gene.score, useNA='ifany')

rm(gene_gene_mat)
```

#### Network analysis

The correlation between eigencentrality and SFARI scores is very small
```{r, fig.width=10}
graph = graph_from_data_frame(edges, vertices=nodes, directed=FALSE)

# Eigencentrality
eigen_centrality_output = eigen_centrality(graph, weights=edges$weight)
eigencentr = eigen_centrality_output$vector

# Strong correlation (negative because 'higher' SFARI scores have actually lower numbers)
cor_SFARI_eigencentr = cor(nodes$gene.score[!is.na(nodes$gene.score)], eigencentr[!is.na(nodes$gene.score)])

plot_data = data.frame('gene.score' = as.character(nodes$gene.score), 'Eigencentrality'=eigencentr)

ggplotly(plot_data %>% ggplot(aes(gene.score, Eigencentrality, fill=gene.score)) + geom_boxplot() + 
         scale_fill_manual(values=SFARI_colour_hue()) + theme_minimal() +
         ggtitle(paste0('Correlation = ', round(cor_SFARI_eigencentr, 4))))

rm(cor_SFARI_eigencentr, plot_data)
```

There doesn't seem to be a strong relation between number of papers and eigencentrality for each score
```{r, fig.width=10, fig.height=6, warning=FALSE}

nodes = nodes %>% mutate('eigencentrality'=eigencentr, 'n.papers'=value, 'gene.score'=as.character(gene.score))

ggplotly(nodes %>% ggplot(aes(n.papers, eigencentrality, fill=gene.score, color=gene.score)) + 
         geom_point(alpha=0.5, aes(id=gene.symbol)) + geom_smooth(method='lm', fill=NA, size=0.5) + 
         theme_minimal() + ggtitle('N papers vs Eigencentrality'))
```

#### Network visualisation
```{r, fig.width=10}
visNetwork(nodes, edges, height='700px', width='100%') %>% visOptions(selectedBy = 'gene.score') %>% 
  visEdges(smooth = TRUE) %>% visIgraphLayout() %>% visLayout(randomSeed=12)
```
