---
title: 'Exploratory Analysis Gupta dataset'
output:
  html_document:
    code_folding: 'hide'
---

```{r, echo=TRUE, include=FALSE}
library(biomaRt)
library(plotly) ; library(plotlyutils) ; require(gridExtra) ; library(viridis)
library(tidyverse)
library(WGCNA)
library(vsn)
library(Rtsne)
library(edgeR)
library(sva)
library(glue)

SFARI_colour_hue = function(r) {
  pal = c('#FF7631','#FFB100','#E8E328','#8CC83F','#62CCA6','#59B9C9','#b3b3b3','#808080','gray','#d9d9d9')[r]
}
```

---

## Gupta normalised dataset (http://www.arkinglab.org/resources/)

**Note:** Raw data can be found in the National Database for Autism Research (NDAR) under the accession code NDARCOL0002034 but you need to request access

* Data preprocessing briefly described in [Transcriptome analysis of cortical tissue reveals shared sets of downregulated genes in autism and schizophrenia](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5070061/)

  * Using conditional quantile normalisation

### Load data

**Note:** The rownames in the expression dataset have the following structure: ensemble ID + '.' + gene name + n. I have to find what the n stands for (just different measurements of the gene? polymorphisms? other?)
```{r load_data, warning=FALSE, message=FALSE}
# LOAD METADATA
datMeta =  read.delim('./../Data/Gupta/AUT_pheno.txt')
datMeta = datMeta %>% dplyr::select(-matches('PC|SV|MEDIAN|.01'))
datMeta$age_group = cut(datMeta$Age, c(-0.5, 0, 0.6, 10, 20, 30, 40, 50, 60, 70, 80),
                    labels=c('Fetal','Infant','Child','10-20','20s','30s','40s','50s','60s','70s'))


# LOAD EXPRESSION DATA
datExpr = read.delim('./../Data/Gupta/NormalizedGeneExpression_AUT.txt')


# ANNOTATE GENES
getinfo = c('ensembl_gene_id','external_gene_id','chromosome_name','start_position',
            'end_position','strand','band','gene_biotype','percentage_gc_content')
mart = useMart(biomart='ENSEMBL_MART_ENSEMBL',
               dataset='hsapiens_gene_ensembl',
               host='feb2014.archive.ensembl.org') ## Gencode v19
datGenes = getBM(attributes=getinfo, filters=c('ensembl_gene_id'), 
                 values=sub('\\..*', '', rownames(datExpr)), mart=mart)
datGenes = datGenes[match(sub('\\..*', '', rownames(datExpr)), datGenes$ensembl_gene_id),]
datGenes$length = datGenes$end_position-datGenes$start_position
datGenes$ID_match_datExpr = paste(datGenes$ensembl_gene_id, gsub('-','.',datGenes$external_gene_id), sep='.')

# SFARI Genes
SFARI_genes = read_csv('./../Data/SFARI/SFARI_genes_with_ensembl_IDs.csv')


# GO Annotations
GO_annotations = read.csv('./../Data/GO_annotations/genes_GO_annotations.csv')
GO_neuronal = GO_annotations %>% filter(grepl('neuron', go_term)) %>% 
              mutate('ID' = as.character(ensembl_gene_id)) %>% 
              dplyr::select(-ensembl_gene_id) %>% distinct(ID) %>%
              mutate('Neuronal' = 1)


# Combine SFARI and GO information
gene_info = data.frame('ID'=sub('\\..*', '', rownames(datExpr))) %>% left_join(SFARI_genes, by='ID') %>% 
            mutate(`gene-score`=ifelse(is.na(`gene-score`), 'None', `gene-score`)) %>%
            left_join(GO_neuronal, by='ID') %>% 
            mutate('gene.score'=ifelse(`gene-score`=='None' & Neuronal==1, 'Neuronal', `gene-score`)) %>%
            mutate('gene.score'=ifelse(is.na(gene.score), 'None', gene.score))


rm(getinfo, mart, GO_annotations)
```
<br>

### Check sample distribution

RNA-Seq for 104 cortical brain-tissue samples across three brain regions (BA10, BA19 and BA44/45), comprising 57 samples from 40 control subjects and 47 samples from 32 ASD subjects

```{r}
print(paste0('Dataset includes ', nrow(datExpr), ' genes from ', ncol(datExpr), ' samples belonging to ', length(unique(datMeta$sampleid)), ' different subjects.'))
```
<br>

**Diagnosis distribution:** Fairly balanced
```{r}
table(datMeta$Dx)
```
<br>

**Brain region distribution:** Most samples (~60%) belong to Brodmann area 19

Where: 1=BA10, 2=BA19 and 3=BA44/45
```{r}
table(datMeta$brainregion)
```
<br>

Diagnosis and brain region are balanced, although not that much
```{r}
table(datMeta$Dx, datMeta$brainregion)
```
<br>

**Sex distribution:** There are three times more Male samples than Female ones
```{r}
table(datMeta$Gender)
```

Fairly balanced
```{r}
table(datMeta$Dx, datMeta$Gender)
```
<br>

**Age distribution:** Subjects between 2 and 82 years old with a mean close to 20
```{r}
summary(datMeta$Age)
```
<br>

---

## Filtering

**Note:** There are some NAs in the expression data, but before diciding which genes/samples to remove it would be better to remove low expressed genes (maybe they have many of the NAs)

```{r}
print(paste0('There are ', sum(is.na(datExpr)), ' NAs in the expression data'))
```

1. Filter genes with low expression levels

Seems like there is a really dense concentration of point with mean expression ~ -1 and sd ~ 1.3
```{r filter_low_expression, fig.width=10}
mean_expr = data.frame('ID'=rownames(datExpr),'Mean'=rowMeans(datExpr, na.rm=T),
                       'SD'=apply(datExpr,1,function(x) sd(x, na.rm=T)))

p1 = mean_expr %>% ggplot(aes(Mean, SD)) + geom_point(alpha=0.01, color='#0099cc') + geom_smooth(method='lm', color='gray') +
     geom_vline(xintercept=-2.8, color='gray', linetype='dashed') + theme_minimal()

p2 = mean_expr %>% ggplot(aes(Mean)) + geom_density(fill='#0099cc', color='#0099cc', alpha=0.5) + 
     geom_vline(xintercept=-2.8, color='gray', linetype='dashed') + theme_minimal()

grid.arrange(p1, p2, ncol=2)

rm(mean_expr, p1, p2)
```

Filtering out genes with a mean expression lower than -2.8
```{r}
to_keep = rowMeans(datExpr, na.rm=TRUE)>-2.8
datExpr = datExpr[to_keep,]

print(paste0('Removed ', sum(!to_keep, na.rm=T), ', ', sum(to_keep, na.rm=T), ' remaining'))

rm(to_keep)
```
<br>

2. Remove NAs in the data

```{r}
missing_points = is.na(datExpr)

print(paste0('There are ', sum(is.na(datExpr)), ' NAs in the expression data'))
```

$\qquad$ 2.1 Filtering samples with the highest number of missing values (chose to remove the top 2 samples, which have over 8 times the average number of missing values per sample)

```{r nas_by_sample, fig.width=10}
nas_by_sample = data.frame('ID' = colnames(datExpr),
                           'NAs' = datExpr %>% apply(2, function(x) x %>% is.na %>% sum))

ggplotly(nas_by_sample %>% ggplot(aes(reorder(ID,-NAs), NAs, fill=NAs)) + geom_bar(stat='identity') + 
         xlab('Sample') + ylab('NA count') + scale_fill_viridis() + theme_minimal() + 
         geom_hline(yintercept=sum(is.na(datExpr))/ncol(datExpr), color='gray') + ggtitle('Missing values by sample') +
         theme(axis.text.x = element_blank(), axis.ticks = element_blank(), legend.position='none'))

print(paste0('Removing samples ', paste(nas_by_sample$ID[nas_by_sample$NAs>1300], collapse = ' and ')))
```

```{r}
to_keep = which(!colnames(datExpr) %in% nas_by_sample$ID[nas_by_sample$NAs>1300])
datExpr = datExpr[,to_keep]
datMeta = datMeta[to_keep,]

print(paste0('Removed ', length(!to_keep), ' samples, ', length(to_keep), ' remaining'))

rm(nas_by_sample, to_keep)
```
<br>

$\qquad$ 2.2 Filter genes with the highest number of missing values (chose to filter the genes with more than 3% missing values (3 or more))
```{r nas_by_gene}
nas_by_gene =  data.frame('ID' = rownames(datExpr),
                           'NAs' = datExpr %>% apply(1, function(x) x %>% is.na %>% sum))

nas_by_gene %>% ggplot(aes(NAs)) + geom_bar(fill='#0099cc') + scale_y_sqrt() + theme_minimal()

to_keep = nas_by_gene$NAs<3
datExpr = datExpr[to_keep,]

print(paste0('Removed ', sum(!to_keep), ' genes, ', sum(to_keep), ' remaining'))

rm(nas_by_gene)
```

$\qquad$ 2.3 Impute missing values using the mice (Multiple Imputations by Chained Equations) package

- If we decide to continue using this dataset we should impute the missing values, but for now I'm just going to remove more samples and genes until there aren't any NAs left

1. Remove all samples with more than 550 missing values

```{r}
to_keep = apply(datExpr, 2, function(x) sum(is.na(x)))<550
datExpr = datExpr[,to_keep]
datMeta = datMeta[to_keep,]

print(paste0('Removed ', sum(!to_keep), ' samples, ', sum(to_keep), ' remaining'))
```

2. Remove all genes with missing values
```{r}
to_keep = apply(datExpr, 1, function(x) sum(is.na(x)))==0
datExpr = datExpr[to_keep,]

print(paste0('Removed ', sum(!to_keep), ' genes, ', sum(to_keep), ' remaining'))
```


<!-- - For reliable results, they recommend that no more than 5% of the values of each variable are missing (our sparsest column has only 2% of the data missing) -->

<!-- ```{r} -->
<!-- mice_output = datExpr[1:1000,] %>% mice(blocks=datMeta$sampleid, printFlag=F, seed=123) -->

<!-- datExpr_imputed = mice_output$ -->
<!-- ``` -->
<br>

3. Filter outlier samples

Using node connectivity as a distance measure, normalising it and filtering out genes farther away than 2 standard deviations from the left (lower connectivity than average, not higher)

- Using $s_{ij}=|bw(i,j)|$ to define connectivity between genes.

- Filtering three samples, all from different subjects but from the Harvard lab

```{r outlier_samples, warning=FALSE}
absadj = datExpr %>% bicor %>% abs
netsummary = fundamentalNetworkConcepts(absadj)
ku = netsummary$Connectivity
z.ku = (ku-mean(ku))/sqrt(var(ku))

plot_data = data.frame('ID'=rownames(datMeta), 'sample'=1:length(z.ku), 'distance'=z.ku, 'Age'=datMeta$age_group)
ggplotly(plot_data %>% ggplot(aes(sample, distance, color=Age)) + geom_point() + geom_hline(yintercept=-2, color='gray') + theme_minimal())

print(paste0('Outlier samples: ', paste(as.character(plot_data$ID[plot_data$distance< -2]), collapse=', ')))
```

```{r}
to_keep = z.ku > -2
datMeta = datMeta[to_keep,]
datExpr = datExpr[,to_keep]

print(paste0('Removed ', sum(!to_keep), ' samples, ', sum(to_keep), ' remaining'))

rm(absadj, netsummary, ku, z.ku, plot_data, to_keep)
```

```{r}
print(paste0('After filtering, the dataset consists of ', nrow(datExpr), ' genes and ', ncol(datExpr), ' samples'))
```
<br><br>

### Differential Expression Analysis

### Batch characterisation

#### !!! ERASED THIS BY MISTAKE
# ```{r}
# table(datMeta$Batch, datMeta$DX)
# ```

### Looking for unknown sources of batch effects with sva

Can't use svaseq because data is not log counts, so using sva instead
```{r sva}
mod = model.matrix(~ Dx, datMeta)
mod0 = model.matrix(~ 1, datMeta)
sva_fit = datExpr %>% as.matrix %>% sva(mod=mod, mod0=mod0)

rm(mod, mod0)
```

Include SV estimations to datMeta information
```{r}
sv_data = sva_fit$sv %>% data.frame
colnames(sv_data) = paste0('SV',1:ncol(sv_data))

datMeta = cbind(datMeta, sv_data)

rm(sv_data)
```
<br><br>

### Differential Expression Analysis

Using lmFit
```{r DEA}
mod = model.matrix(~ SV1 + SV2 + SV3 + SV4 + SV5 + Dx, data=datMeta)
corfit = duplicateCorrelation(datExpr, mod, block=datMeta$sampleid)
lmfit = lmFit(datExpr, design=mod, block=datMeta$sampleid, correlation=corfit$consensus)

fit = eBayes(lmfit, trend=T, robust=T)
top_genes = topTable(fit, coef=2, number=nrow(datExpr)) %>% mutate('ID'=sub('\\..*', '', rownames(.)))

DE_info = gene_info %>% left_join(top_genes, by='ID')

rm(mod, corfit, lmfit, fit, top_genes)
```
<br><br>

### Visualisations
<br>

#### Samples

**PCA:** There doesn't seem to be a specific pattern related to gender or age, there does seem to be a small differentiation by diagnosis in the 2nd principal component

```{r pca_samples, warning=FALSE, message=FALSE, fig.width=10}
pca = datExpr %>% t %>% prcomp

plot_data = data.frame('ID'=colnames(datExpr), 'PC1' = pca$x[,1], 'PC2' = pca$x[,2]) %>% 
            left_join(datMeta, by='ID') %>% 
            dplyr::select('PC1','PC2','Age','age_group','Gender','Dx') %>%
            mutate('sqrt_age' = sqrt(Age))

selectable_scatter_plot(plot_data, plot_data[,-c(1,2)])

rm(pca, plot_data)
```
<br>

#### Genes

- First Principal Component explains 92% of the total variance

- There's a really strong (negative) correlation between the mean expression of a gene and the 1st principal component

- It seems like the lower expressed points that make a weird turn on the PC2 axis have a smaller PC3 value than the rest (rotate PC1,PC2 axis and it becomes visible). Haven't found what else characterises them

```{r pca_genes, fig.width=10, warning=FALSE, message=FALSE}
pca = datExpr %>% prcomp

plot_data = data.frame('PC1' = pca$x[,1], 'PC2' = pca$x[,2], 'PC3' = pca$x[,3], 
                       'MeanExpr'=rowMeans(datExpr), 'SD'=apply(datExpr,1,sd))

plot_data %>% ggplot(aes(PC1, PC2, color=MeanExpr)) + geom_point(alpha=0.2) + theme_minimal() + 
     scale_color_viridis() + ggtitle('PCA') +
     xlab(paste0('PC1 (',round(100*summary(pca)$importance[2,1],1),'%)')) +
     ylab(paste0('PC2 (',round(100*summary(pca)$importance[2,2],1),'%)'))

plot_data %>% plot_ly(x=~PC1, y=~PC2, z=~PC3) %>% add_markers(color=plot_data$MeanExpr, size=1) %>%
    layout(title = 'PCA 3D',
           scene = list(xaxis=list(title=glue('PC1 (',round(summary(pca)$importance[2,1]*100,2),'%)')),
                        yaxis=list(title=glue('PC2 (',round(summary(pca)$importance[2,2]*100,2),'%)')),
                        zaxis=list(title=glue('PC3 (',round(summary(pca)$importance[2,3]*100,2),'%)'))))

rm(pca, plot_data)
```
<br><br>

---

## SFARI scores

### Mean and SD

**The higher the SFARI score the higher the mean expression but the lower the standard deviation**. Same as Gandal's dataset!
```{r SFARI_score_LFC, warning=FALSE, fig.width=10}
plot_data = data.frame('ID'=sub('\\..*', '', rownames(datExpr)),
                       'MeanExpr'=rowMeans(datExpr), 'SDExpr'=apply(datExpr,1,sd)) %>% 
            left_join(DE_info, by='ID')

p1 = ggplotly(plot_data %>% ggplot(aes(gene.score, MeanExpr, fill=gene.score)) + geom_boxplot() + 
              scale_fill_manual(values=SFARI_colour_hue(r=c(1:6,8,7))) + theme_minimal() +
              theme(legend.position='none'))

p2 = ggplotly(plot_data %>% ggplot(aes(gene.score, SDExpr, fill=gene.score)) + geom_boxplot() + 
              scale_fill_manual(values=SFARI_colour_hue(r=c(1:6,8,7))) + theme_minimal() +
              ggtitle('Mean Expression (left) and SD (right) by SFARI score') + 
              theme(legend.position='none'))

subplot(p1, p2, nrows=1)

rm(plot_data, p1, p2)
```
<br>

### Log Fold Change

**From scores 1 through 4, the higher the SFARI score, the lower the log Fold Change**. Same behaviour as Gandal!
```{r}
ggplotly(DE_info %>% ggplot(aes(x=gene.score, y=abs(logFC), fill=gene.score)) + 
         geom_boxplot() + scale_fill_manual(values=SFARI_colour_hue(r=c(1:6,8,7))) + 
         theme_minimal() + theme(legend.position='none'))
```
<br><br>

---

#### Save preprocessed dataset
```{r save_preprocessed_dataset}
save(datExpr, datMeta, datGenes, DE_info, file='./../Data/Gupta/preprocessed_data.RData')
#load('./../Data/Gupta/preprocessed_data.RData')
```
<br>

---

#### Session info
```{r session_info}
sessionInfo()
```
<br><br>
