---
title: 'Simulated Data Exploratory Analysis'
output:
  html_document:
    code_folding: 'hide'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse) ; library(plotly) ; library(reshape2)
library(DESeq2)

load('./../Data/Gandal_RNASeq.RData')
```

### Problem with $log_2$ transformation

Can underestimate variance of genes with a high level of expression

**Experiment:** Generate data with different properties to see how the transformation affects the variance

#### Generating the dataset

The gene expression grouped by SFARI could be approximated by a logNormal distribution ... kind of...

```{r, warning=FALSE}
all_points = datExpr %>% mutate('ID' = rownames(datExpr)) %>% left_join(SFARI_genes, by='ID') %>%
            filter(!is.na(`gene-score`)) %>% mutate(`gene-score` = as.factor(`gene-score` )) %>% 
            dplyr::select(c(colnames(datExpr), `gene-score`)) %>% melt

ggplotly(ggplot(all_points, aes(value, group=`gene-score`, fill=`gene-score`, color=`gene-score`)) + 
         geom_density(alpha=0.3, color=NA) + scale_x_continuous(trans='log2') + theme_minimal() +
         ggtitle('Gene expression grouped by SFARI score'))
```

```{r}
generate_logNormal_samples = function(SFARI_score, m, b){
  datExpr_score = datExpr %>% filter(rownames(datExpr) %in% SFARI_genes$ID[SFARI_genes$`gene-score`==SFARI_score])
  score_row_means = log2(datExpr_score+1) %>% rowMeans
  normal_samples = rnorm(nrow(datExpr_score), mean(score_row_means), sd(score_row_means))
  logNormal_samples = data.frame('' ,''=2^normal_samples)

  return(logNormal_samples)
}

datExpr_mean_var = data.frame('ID' = rownames(datExpr), 'mean'=rowMeans(datExpr), 'sd'=apply(datExpr,1,sd))
linReg = lm(sd~mean, data=datExpr_mean_var)
intercept = linReg$coefficients[[1]]
slope = linReg$coefficients[[2]]

print(paste0('The linear regression between mean and standard deviation for the original data has an intercept of ', intercept, ' and a slope of ', slope))


for(score in seq(1,6)){
  
}

rm(linReg, intercept, slope)
```


### Problem with logFC

**"A disadvantage and serious risk of using fold change in this setting is that it is biased and may misclassify differentially expressed genes with large differences (B − A) but small ratios (B/A), leading to poor identification of changes at high expression levels"**. This is a big problem with the SFARI genes because they have very high expression levels, especially the highest scores, so they are the most affected by this.

**Experiment:** Translate the dataset 10 units at a time and see how the logFC values change

- Used only SFARI genes to make the calculations faster

- The adjusted p-value is not affected by the gene expression of the dataset but the logFC is. The bigger the translation, the smaller the number of genes with a significant logFC

```{r, warning=FALSE, message=FALSE, fig.width=10}
get_logFC = function(datExpr, trans){
  
  datExpr_SFARI = datExpr %>% filter(rownames(datExpr) %in% SFARI_genes$ID)
  rownames(datExpr_SFARI) = rownames(datExpr)[rownames(datExpr) %in% SFARI_genes$ID]
  datExpr_translated = datExpr_SFARI + 10*trans
  datProbes_SFARI = datProbes %>% filter(rownames(datExpr) %in% SFARI_genes$ID)
  counts = as.matrix(datExpr_translated)
  rowRanges = GRanges(datProbes_SFARI$chromosome_name,
                      IRanges(datProbes_SFARI$start_position, width=datProbes_SFARI$length),
                      strand=datProbes_SFARI$strand,
                      feature_id=datProbes_SFARI$ensembl_gene_id)
  
  se = SummarizedExperiment(assays=SimpleList(counts=counts), rowRanges=rowRanges, colData=datMeta)
  ddsSE = DESeqDataSet(se, design =~Diagnosis_)
  
  dds = DESeq(ddsSE)
  DE_info = results(dds) %>% data.frame %>% rownames_to_column(var = 'ID') %>%
            mutate('logFC'=log2FoldChange, 'translation'=10*trans) %>% 
            dplyr::select(ID, logFC, padj, translation)
  
  return(DE_info)
}

logFC_results = get_logFC(datExpr, 0)

for(trans in seq(1,10)){
  new = get_logFC(datExpr, trans)
  logFC_results = bind_rows(logFC_results, new)
}

p1 = ggplotly(ggplot(logFC_results, aes(translation, abs(logFC), group=translation, fill=translation)) + 
              geom_boxplot() + theme_minimal())

p2 = ggplotly(ggplot(logFC_results, aes(translation, padj, group=translation, fill=translation)) + 
              geom_boxplot() + theme_minimal() + 
              ggtitle('LogFC (left) and adj p-val (right) for different translations of the data'))

subplot(p1, p2, nrows=1)

significant = logFC_results %>% filter(abs(logFC)>log2(1.2) & padj<0.05) %>%  group_by(translation) %>% tally

ggplotly(ggplot(significant, aes(translation, n, fill=translation)) + geom_bar(stat='identity') +
         theme_minimal() + ggtitle('Number of DE genes for each translation'))

rm(new, p1, p2, logFC_results, significant)
```

### Replicate the behaviour for the different SFARI scores

The gene expression grouped by SFARI could be approximated by a logNormal distribution ... kind of...

```{r, warning=FALSE}
all_points = datExpr %>% mutate('ID' = rownames(datExpr)) %>% left_join(SFARI_genes, by='ID') %>%
            filter(!is.na(`gene-score`)) %>% mutate(`gene-score` = as.factor(`gene-score` )) %>% 
            dplyr::select(c(colnames(datExpr), `gene-score`)) %>% melt

ggplotly(ggplot(all_points, aes(value, group=`gene-score`, fill=`gene-score`, color=`gene-score`)) + 
         geom_density(alpha=0.3, color=NA) + scale_x_continuous(trans='log2') + theme_minimal() +
         ggtitle('Gene expression grouped by SFARI score'))

```

<!-- Using just the means of each row instead of the whole information seems to preserve the structure seen above and it makes the distributions a bit smoother. -->
<!-- ```{r, warning=FALSE} -->
<!-- row_means = data.frame('ID' = rownames(datExpr), 'row_means'=rowMeans(datExpr)) %>%  -->
<!--             left_join(SFARI_genes, by='ID') %>% dplyr::select(ID, row_means, `gene-score`) %>% -->
<!--             filter(!is.na(`gene-score`)) %>% mutate(`gene-score` = as.factor(`gene-score` )) -->

<!-- ggplotly(ggplot(row_means, aes(row_means, group=`gene-score`, fill=`gene-score`, color=`gene-score`)) +  -->
<!--          geom_density(alpha=0.3, color=NA) + scale_x_continuous(trans='log2') + theme_minimal() + -->
<!--          ggtitle('Mean gene expression grouped by SFARI score')) -->

<!-- table(row_means$`gene-score`) -->
<!-- ``` -->
<!-- Both groups of Diagnosis could be approximated by a logNormal distribution for each SFARI score as well -->
<!-- ```{r, warning=FALSE} -->
<!-- ASD_CTL_points = datMeta %>% mutate('variable'=paste0('X',datMeta$Dissected_Sample_ID)) %>% -->
<!--                  dplyr::select(variable, Diagnosis_) %>% right_join (all_points, by='variable') -->

<!-- ggplotly(ggplot(ASD_CTL_points, aes(value, group=Diagnosis_, fill=Diagnosis_, color=Diagnosis_)) +  -->
<!--          geom_density(alpha=0.3, color=NA) + scale_x_continuous(trans='log2') + theme_minimal() + -->
<!--          ggtitle('Gene expression grouped by Diagnosis for each SFARI score') + facet_wrap(~`gene-score`, ncol=2)) -->
<!-- ``` -->

<!-- #### Generate artificial distributions for ASD and CTL for each SFARI score -->

<!-- - Transform the data into logarithmic scale -->
<!-- - Calculate the mean and variance of each Diagnosis and SFARI score group -->
<!-- - Generate a new dataset with random samples selected from these distributions -->

<!-- ```{r} -->
<!-- log2(datExpr+1) -->
<!-- ``` -->





