---
title: 'Chow Exploratory Analysis'
output:
  html_document:
    code_folding: 'hide'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=TRUE, include=FALSE}
#setwd('/afs/inf.ed.ac.uk/user/s17/s1725186/Documents/PhD-InitialExperiments/SFARI_genes/R_markdowns')

library(tidyverse) ; library(reshape2) ; library(glue) ; library(plotly)
library(RColorBrewer) ; library(viridis) ; require(gridExtra) ; library(plotlyutils)
library(biomaRt) ; library(lumi)
library(AnnotationDbi) ; library(illuminaHumanv4.db) ; library(samr)

gg_colour_hue = function(n) {
  hues = seq(15, 375, length = n-2)
  pal = c(hcl(h = hues, l = 65, c = 100)[1:(n-3)], '#666666', '#8c8c8c', '#bfbfbf')
}
```

## Normalised dataset
```{r, warning=FALSE, message=FALSE}

if(!file.exists('./../Data/Chow_normalised.RData')){
  
  # Download and load Microarray_ASD_Chow_normalized.RData
  datMeta$Disease.status = datMeta$DX
  
  # Perform Differential Expression Analysis
  SAM_fit = SAM(as.matrix(datExpr), as.factor(datMeta$Disease.status), resp.type = 'Two class unpaired', 
                geneid = rownames(datExpr), random.seed = 1234, logged2 = TRUE, fdr.output = 1)
  genes_up = SAM_fit$siggenes.table$genes.up %>% data.frame
  genes_down = SAM_fit$siggenes.table$genes.lo %>% data.frame
  DE_info = genes_up %>% bind_rows(genes_down)
  
  # Load SFARI information
  SFARI_genes = read_csv('./../../Gandal/RNAseq/working_data/SFARI_genes_01-15-2019.csv')
  ilmn_gene_map = mapIds(illuminaHumanv4.db, key=rownames(datExpr), 
                         column=c('SYMBOL'), keytype='PROBEID', multiVals='filter')
  SFARI_genes = data.frame('ID'=names(ilmn_gene_map), 'gene-symbol'=unname(ilmn_gene_map)) %>%
                 right_join(SFARI_genes, by=c('gene.symbol'='gene-symbol'))
  
  # Add neuronal annotations
  GO_annotations = read.csv('./../../Gandal/RNAseq/working_data/genes_GO_annotations.csv')
  GO_neuronal = GO_annotations %>% filter(grepl('neuron', go_term)) %>% 
                distinct(ensembl_gene_id) %>%
                mutate('Neuronal' = 1)
  
  mart = useMart(biomart='ENSEMBL_MART_ENSEMBL', dataset='hsapiens_gene_ensembl',
                 host='feb2014.archive.ensembl.org')
  ensembl_gene_map = getBM(attributes=c('ensembl_gene_id', 'hgnc_symbol'), filters=c('ensembl_gene_id'), 
                     values=GO_neuronal$ensembl_gene_id, mart=mart)
  ensembl_ilmn_gene_map = data.frame('ID' = names(ilmn_gene_map), 'gene.symbol'=unname(ilmn_gene_map)) %>% 
                          full_join(ensembl_gene_map, by=c('gene.symbol'='hgnc_symbol'))
  
  GO_neuronal = GO_neuronal %>% left_join(ensembl_ilmn_gene_map, by='ensembl_gene_id') %>%
                filter(!is.na(ID))
  
  save(datExpr, datMeta, GO_neuronal, SFARI_genes, DE_info, file='./../Data/Chow_normalised.RData')
  
} else load('./../Data/Chow_normalised.RData')

datExpr_backup = datExpr

DE_info = DE_info %>% mutate('logFC'= Fold.Change, 'ID' = Gene.Name)

datMeta = datMeta %>% mutate('Diagnosis_' = ifelse(DX=='Autism','autism','control'))

```

Number of genes and samples:
```{r}
dim(datExpr)
```
<br>

Gene count by SFARI score of remaining genes:
```{r}
table(SFARI_genes$`gene-score`)
```
<br>

Relation between SFARI scores and Neuronal functional annotation:

```{r, warning=FALSE}
Neuronal_SFARI = data.frame('ID'=rownames(datExpr), 'Neuronal'=rownames(datExpr) %in% GO_neuronal$ID) %>%
                 left_join(SFARI_genes, by='ID')

tbl = table(Neuronal_SFARI$`gene-score`, Neuronal_SFARI$Neuronal, useNA='ifany') %>% t %>% as.data.frame.matrix
rownames(tbl) = c('Non-Neuronal','Neuronal')
tbl

tbl = round(sweep(tbl, 2, colSums(tbl), `/`)*100, 2)
tbl

rm(Neuronal_SFARI, tbl)
```

<br><br>

### Boxplots

#### Mean and Standard Deviation by score

```{r, fig.width=10, warning=FALSE}

make_ASD_vs_CTL_df = function(datExpr){
  datExpr_ASD = datExpr %>% data.frame %>% dplyr::select(which(datMeta$Diagnosis_=='autism'))
  datExpr_CTL = datExpr %>% data.frame %>% dplyr::select(which(datMeta$Diagnosis_=='control'))
  
  ASD_vs_CTL = data.frame('ID'=as.character(rownames(datExpr)),
                          'mean' = rowMeans(datExpr), 'sd' = apply(datExpr, 1, sd),
                          'mean_ASD'=rowMeans(datExpr_ASD), 'mean_CTL'=rowMeans(datExpr_CTL),
                          'sd_ASD'=apply(datExpr_ASD,1,sd), 'sd_CTL'=apply(datExpr_CTL,1,sd)) %>%
               mutate('mean_diff'=mean_ASD-mean_CTL, 'sd_diff'=sd_ASD-sd_CTL) %>%
               left_join(SFARI_genes, by='ID') %>%
               dplyr::select(ID, mean, sd, mean_ASD, mean_CTL, mean_diff, sd_ASD, sd_CTL, sd_diff, `gene-score`) %>%
               mutate('Neuronal'=ifelse(ID %in% GO_neuronal$ID, 'Neuronal', 'Non-Neuronal'))
  
  ASD_vs_CTL_SFARI = ASD_vs_CTL %>% filter(!is.na(`gene-score`)) %>% 
                     mutate(Group = as.character(`gene-score`)) %>% dplyr::select(-c(Neuronal,`gene-score`))
  ASD_vs_CTL_Neuro = ASD_vs_CTL %>% mutate(Group = Neuronal) %>% dplyr::select(-c(Neuronal,`gene-score`))
  ASD_vs_CTL_all = ASD_vs_CTL %>% mutate(Group = 'All') %>% dplyr::select(-c(Neuronal,`gene-score`))
  
  ASD_vs_CTL_together = bind_rows(ASD_vs_CTL_SFARI, ASD_vs_CTL_Neuro, ASD_vs_CTL_all) %>% 
                        mutate(Group = ordered(Group, levels=c('1','2','3','4','5','6',
                                                               'Neuronal','Non-Neuronal','All'))) %>%
                        left_join(DE_info, by='ID')
  
  return(ASD_vs_CTL_together)
}

ASD_vs_CTL = make_ASD_vs_CTL_df(datExpr)

p1 = ggplotly(ggplot(ASD_vs_CTL, aes(Group, mean, fill=Group)) + 
              geom_boxplot() + theme_minimal() + 
              ggtitle('Mean (left) and Standard Deviation (right) by score') +
              scale_fill_manual(values=gg_colour_hue(9)) +
              theme(legend.position='none', axis.text.x=element_text(angle=45, hjust=1)))

p2 = ggplotly(ggplot(ASD_vs_CTL, aes(Group, sd, fill=Group)) + 
              geom_boxplot() + theme_minimal() +
              ggtitle('Mean (left) and Standard Deviation (right) by score') +
              scale_fill_manual(values=gg_colour_hue(9)) +
              theme(legend.position='none', axis.text.x=element_text(angle=45, hjust=1)))

subplot(p1, p2, nrows=1)

rm(p1, p2)
```

The relation doesn't seem to be completely linear, but close to.
```{r, warning=FALSE, fig.width=10}
ggplot(ASD_vs_CTL, aes(mean, sd)) + geom_point(alpha=0.1, aes(id=ID, fill=Group, color=Group)) +
         scale_color_manual(values=gg_colour_hue(9)) + scale_fill_manual(values=gg_colour_hue(9)) +  
         scale_x_continuous(trans='log2') + scale_y_continuous(trans='log2') + 
           ggtitle(paste0('Corr=',round(cor(ASD_vs_CTL$mean, ASD_vs_CTL$sd), 2))) +
         geom_smooth(method=lm, se=FALSE, color='#808080', size=0.5) + theme_minimal()
```

#### Difference between Diagnosis by score

```{r, fig.width=10, warning=FALSE}
p1 = ggplotly(ggplot(ASD_vs_CTL, aes(Group, abs(mean_diff), fill=Group)) + 
                geom_boxplot() + theme_minimal() + 
                ggtitle('Difference in mean (left) and log Fold Change (right) by Diagnosis') +
                scale_fill_manual(values=gg_colour_hue(9)) +
                theme(legend.position='none', axis.text.x=element_text(angle=45, hjust=1)))

p2 = ggplotly(ggplot(ASD_vs_CTL[complete.cases(ASD_vs_CTL),], aes(Group, abs(as.numeric(logFC)), fill=Group)) + 
                geom_boxplot() + theme_minimal() +
                ggtitle('Difference in mean (left) and log Fold Change (right) by Diagnosis') +
                scale_fill_manual(values=gg_colour_hue(9)) +
                theme(legend.position='none', axis.text.x=element_text(angle=45, hjust=1)))

subplot(p1, p2, nrows=1)

rm(p1, p2)
```

## Batch corrected dataset
```{r, warning=FALSE, message=FALSE}

if(!file.exists('./../Data/Chow_combat.RData')){
  
  # Download and load Microarray_ASD_chow_normalized_CR_cleaned
  datMeta$Disease.status = datMeta$DX
  
  # Perform Differential Expression Analysis
  SAM_fit = SAM(as.matrix(datExpr), as.factor(datMeta$Disease.status), resp.type = 'Two class unpaired', 
                geneid = rownames(datExpr), random.seed = 1234, logged2 = TRUE, fdr.output = 1)
  genes_up = SAM_fit$siggenes.table$genes.up %>% data.frame
  genes_down = SAM_fit$siggenes.table$genes.lo %>% data.frame
  DE_info = genes_up %>% bind_rows(genes_down)
  
  # Load SFARI information
  SFARI_genes = read_csv('./../../Gandal/RNAseq/working_data/SFARI_genes_01-15-2019.csv')
  
  mart = useMart(biomart='ENSEMBL_MART_ENSEMBL',
                 dataset='hsapiens_gene_ensembl',
                 host='feb2014.archive.ensembl.org') ## Gencode v19
  
  gene_names = getBM(attributes=c('ensembl_gene_id', 'hgnc_symbol'), filters=c('hgnc_symbol'), 
                     values=SFARI_genes$`gene-symbol`, mart=mart) %>% 
                     mutate('gene-symbol'=hgnc_symbol, 'ID'=as.character(ensembl_gene_id)) %>% 
                     dplyr::select('ID', 'gene-symbol')
  
  SFARI_genes = left_join(SFARI_genes, gene_names, by='gene-symbol')
  
  # Add neuronal annotations
  GO_annotations = read.csv('./../../Gandal/RNAseq/working_data/genes_GO_annotations.csv')
  GO_neuronal = GO_annotations %>% filter(grepl('neuron', go_term)) %>% 
                distinct(ensembl_gene_id) %>%
                mutate('Neuronal' = 1, 'ID'=ensembl_gene_id)
  
  save(datExpr, datMeta, GO_neuronal, SFARI_genes, DE_info, file='./../Data/Chow_combat.RData')
  
} else load('./../Data/Chow_combat.RData')

datExpr_backup = datExpr

DE_info = DE_info %>% mutate('logFC'= Fold.Change, 'ID' = Gene.Name)

datMeta = datMeta %>% mutate('Diagnosis_' = ifelse(DX=='Autism','autism','control'))

```

Number of genes and samples:
```{r}
dim(datExpr)
```
<br>

Gene count by SFARI score of remaining genes:
```{r}
table(SFARI_genes$`gene-score`)
```
<br>

Relation between SFARI scores and Neuronal functional annotation:

```{r, warning=FALSE}
Neuronal_SFARI = data.frame('ID'=rownames(datExpr), 'Neuronal'=rownames(datExpr) %in% GO_neuronal$ID) %>%
                 left_join(SFARI_genes, by='ID')

tbl = table(Neuronal_SFARI$`gene-score`, Neuronal_SFARI$Neuronal, useNA='ifany') %>% t %>% as.data.frame.matrix
rownames(tbl) = c('Non-Neuronal','Neuronal')
tbl

tbl = round(sweep(tbl, 2, colSums(tbl), `/`)*100, 2)
tbl

rm(Neuronal_SFARI, tbl)
```

<br><br>

### Boxplots

#### Mean and Standard Deviation by score

```{r, fig.width=10, warning=FALSE}

make_ASD_vs_CTL_df = function(datExpr){
  datExpr_ASD = datExpr %>% data.frame %>% dplyr::select(which(datMeta$Diagnosis_=='autism'))
  datExpr_CTL = datExpr %>% data.frame %>% dplyr::select(which(datMeta$Diagnosis_=='control'))
  
  ASD_vs_CTL = data.frame('ID'=as.character(rownames(datExpr)),
                          'mean' = rowMeans(datExpr), 'sd' = apply(datExpr, 1, sd),
                          'mean_ASD'=rowMeans(datExpr_ASD), 'mean_CTL'=rowMeans(datExpr_CTL),
                          'sd_ASD'=apply(datExpr_ASD,1,sd), 'sd_CTL'=apply(datExpr_CTL,1,sd)) %>%
               mutate('mean_diff'=mean_ASD-mean_CTL, 'sd_diff'=sd_ASD-sd_CTL) %>%
               left_join(SFARI_genes, by='ID') %>%
               dplyr::select(ID, mean, sd, mean_ASD, mean_CTL, mean_diff, sd_ASD, sd_CTL, sd_diff, `gene-score`) %>%
               mutate('Neuronal'=ifelse(ID %in% GO_neuronal$ID, 'Neuronal', 'Non-Neuronal'))
  
  ASD_vs_CTL_SFARI = ASD_vs_CTL %>% filter(!is.na(`gene-score`)) %>% 
                     mutate(Group = as.character(`gene-score`)) %>% dplyr::select(-c(Neuronal,`gene-score`))
  ASD_vs_CTL_Neuro = ASD_vs_CTL %>% mutate(Group = Neuronal) %>% dplyr::select(-c(Neuronal,`gene-score`))
  ASD_vs_CTL_all = ASD_vs_CTL %>% mutate(Group = 'All') %>% dplyr::select(-c(Neuronal,`gene-score`))
  
  ASD_vs_CTL_together = bind_rows(ASD_vs_CTL_SFARI, ASD_vs_CTL_Neuro, ASD_vs_CTL_all) %>% 
                        mutate(Group = ordered(Group, levels=c('1','2','3','4','5','6',
                                                               'Neuronal','Non-Neuronal','All'))) %>%
                        left_join(DE_info, by='ID')
  
  return(ASD_vs_CTL_together)
}

ASD_vs_CTL = make_ASD_vs_CTL_df(datExpr)

p1 = ggplotly(ggplot(ASD_vs_CTL, aes(Group, mean, fill=Group)) + 
              geom_boxplot() + theme_minimal() + 
              ggtitle('Mean (left) and Standard Deviation (right) by score') +
              scale_fill_manual(values=gg_colour_hue(9)) +
              theme(legend.position='none', axis.text.x=element_text(angle=45, hjust=1)))

p2 = ggplotly(ggplot(ASD_vs_CTL, aes(Group, sd, fill=Group)) + 
              geom_boxplot() + theme_minimal() +
              ggtitle('Mean (left) and Standard Deviation (right) by score') +
              scale_fill_manual(values=gg_colour_hue(9)) +
              theme(legend.position='none', axis.text.x=element_text(angle=45, hjust=1)))

subplot(p1, p2, nrows=1)

rm(p1, p2)
```

```{r, warning=FALSE, fig.width=10}
ggplot(ASD_vs_CTL, aes(mean, sd)) + geom_point(alpha=0.1, aes(id=ID, fill=Group, color=Group)) +
         scale_color_manual(values=gg_colour_hue(9)) + scale_fill_manual(values=gg_colour_hue(9)) +  
         scale_x_continuous(trans='log2') + scale_y_continuous(trans='log2') + 
           ggtitle(paste0('Corr=',round(cor(ASD_vs_CTL$mean, ASD_vs_CTL$sd), 2))) +
         geom_smooth(method=lm, se=FALSE, color='#808080', size=0.5) + theme_minimal()
```

#### Difference between Diagnosis by score

```{r, fig.width=10, warning=FALSE}
p1 = ggplotly(ggplot(ASD_vs_CTL, aes(Group, abs(mean_diff), fill=Group)) + 
                geom_boxplot() + theme_minimal() + 
                ggtitle('Difference in mean (left) and log Fold Change (right) by Diagnosis') +
                scale_fill_manual(values=gg_colour_hue(9)) +
                theme(legend.position='none', axis.text.x=element_text(angle=45, hjust=1)))

p2 = ggplotly(ggplot(ASD_vs_CTL[complete.cases(ASD_vs_CTL),], aes(Group, abs(as.numeric(logFC)), fill=Group)) + 
                geom_boxplot() + theme_minimal() +
                ggtitle('Difference in mean (left) and log Fold Change (right) by Diagnosis') +
                scale_fill_manual(values=gg_colour_hue(9)) +
                theme(legend.position='none', axis.text.x=element_text(angle=45, hjust=1)))

subplot(p1, p2, nrows=1)

rm(p1, p2)
```




