---
title: 'BrainSpan Exploratory Analysis'
output:
  html_document:
    code_folding: 'hide'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=TRUE, include=FALSE}
#setwd('/afs/inf.ed.ac.uk/user/s17/s1725186/Documents/PhD-InitialExperiments/SFARI_genes/R_markdowns')

library(tidyverse) ; library(reshape2) ; library(glue) ; library(plotly)
library(RColorBrewer) ; library(viridis) ; require(gridExtra) ; library(plotlyutils)
library(biomaRt) ; #library(DESeq2) ; library(limma) ; library(cqn)
library(venneuler)
library(limma) ; library(edgeR)

gg_colour_hue = function(n) {
  hues = seq(15, 375, length = n-2)
  pal = c(hcl(h = hues, l = 65, c = 100)[1:(n-3)], '#666666', '#8c8c8c', '#bfbfbf')
}
```

```{r, warning=FALSE, message=FALSE}
if(!file.exists('./../Data/BrainSpan_raw.RData')){
  
  # Load csvs (Downloaded from 'RNA-Seq Gencode v10 summarized to genes' in http://www.brainspan.org/static/download.html)
  datExpr = read.csv('./../Data/BrainSpan_genes/expression_matrix.csv', header=FALSE)
  datMeta = read.csv('./../Data/BrainSpan_genes/columns_metadata.csv')
  geneInfo = read.csv('./../Data/BrainSpan_genes/rows_metadata.csv')

  # Remove index column in datExpr
  cols = datExpr %>% colnames
  datExpr = datExpr %>% dplyr::select(-V1)
  colnames(datExpr) = cols[-length(cols)]
  
  # Make sure rows match
  if(!all(rownames(datExpr) == geneInfo$row_num)){
   print('Columns in datExpr don\'t match the rows in datMeta!') 
  }
  
  # Assign row names
  rownames(datMeta) = paste0('V', datMeta$column_num)
  rownames(datExpr) = geneInfo$ensembl_gene_id
  
  # Annotate probes
  getinfo = c('ensembl_gene_id','external_gene_id','chromosome_name','start_position',
              'end_position','strand','band','gene_biotype','percentage_gc_content')
  mart = useMart(biomart='ENSEMBL_MART_ENSEMBL',
                 dataset='hsapiens_gene_ensembl',
                 host='feb2014.archive.ensembl.org') ## Gencode v19
  datProbes = getBM(attributes=getinfo, filters=c('ensembl_gene_id'), values=rownames(datExpr), mart=mart)
  datProbes = datProbes[match(rownames(datExpr), datProbes$ensembl_gene_id),]
  datProbes$length = datProbes$end_position-datProbes$start_position
  
  # Group brain regions by lobes
  datMeta$Brain_Region = as.factor(datMeta$structure_acronym)
  datMeta$Brain_lobe = 'Other'
  datMeta$Brain_lobe[datMeta$Brain_Region %in% c('Ocx','V1C')] = 'Occipital'
  datMeta$Brain_lobe[datMeta$Brain_Region %in% c('M1C-S1C','DFC','OFC','VFC','M1C')] = 'Frontal'
  datMeta$Brain_lobe[datMeta$Brain_Region %in% c('PCx','IPC', 'S1C')] = 'Parietal'
  datMeta$Brain_lobe[datMeta$Brain_Region %in% c('AMY','MGE','STC','ITC','HIP','TCx','A1C')] = 'Temporal'
  datMeta$Brain_lobe[datMeta$Brain_Region %in% c('MFC')] = 'Limbic'
  datMeta$Brain_lobe=factor(datMeta$Brain_lobe, levels=c('Frontal', 'Temporal', 'Parietal', 'Occipital','Limbic','Other'))
  # Don't correspond to any lobe: URL, DTH, LGE, STR, CB, CBC, MD, CGE

  #################################################################################
  # INITIAL FILTERING:
  
  # 1 Filter probes with start or end position missing (none)
  to_keep = !is.na(datProbes$length)
  datProbes = datProbes[to_keep,]
  datExpr = datExpr[to_keep,]
  rownames(datProbes) = datProbes$ensembl_gene_id
  
  # 2. Keep only Frontal and Temporal samples to do DEA (filter 301 samples)
  to_keep = datMeta$Brain_lobe %in% c('Temporal','Frontal')
  datExpr = datExpr[,to_keep]
  datMeta = datMeta[to_keep,]
  datMeta$Brain_lobe = factor(datMeta$Brain_lobe, levels=c('Temporal','Frontal'))
  
  # 3. Filter probes with only zeros as entries (filter 2056 probes)
  to_keep = rowSums(datExpr)>0
  datProbes = datProbes[to_keep,]
  datExpr = datExpr[to_keep,]
  
  #################################################################################
  # Annotate SFARI genes
  
  SFARI_genes = read_csv('./../../Gandal/RNAseq/working_data/SFARI_genes_01-15-2019.csv')
  
  # Get ensemble IDS for SFARI genes
  mart = useMart(biomart='ENSEMBL_MART_ENSEMBL',
                 dataset='hsapiens_gene_ensembl',
                 host='feb2014.archive.ensembl.org') ## Gencode v19
  
  gene_names = getBM(attributes=c('ensembl_gene_id', 'hgnc_symbol'), filters=c('hgnc_symbol'), 
                     values=SFARI_genes$`gene-symbol`, mart=mart) %>% 
                     mutate('gene-symbol'=hgnc_symbol, 'ID'=as.character(ensembl_gene_id)) %>% 
                     dplyr::select('ID', 'gene-symbol')
  
  SFARI_genes = left_join(SFARI_genes, gene_names, by='gene-symbol')
  
  #################################################################################
  # Add functional annotation to genes from GO
  
  GO_annotations = read.csv('./../../Gandal/RNAseq/working_data/genes_GO_annotations.csv')
  GO_neuronal = GO_annotations %>% filter(grepl('neuron', go_term)) %>% 
                mutate('ID' = as.character(ensembl_gene_id)) %>% 
                dplyr::select(-ensembl_gene_id) %>% distinct(ID) %>%
                mutate('Neuronal' = 1)
  
  save(datExpr, datMeta, datProbes, SFARI_genes, GO_neuronal, file='./../Data/BrainSpan_raw.RData')
  
  rm(gene_names, geneInfo, GO_annotations, mart, cols, getinfo, to_keep)
  
} else load('./../Data/BrainSpan_raw.RData')
```

### Visualisations

#### PCA

- MDS takes too long to calculate distances

- There doesn't seem to be a clear pattern for SFARI scores

- Applied log2 transformation to the data before performing pca to help the visualisation: **% variance explained by 1PC changed from 60 to 90%**. This was probably what was happening with Gandal's dataset as well

```{r, warning=FALSE}
datExpr_pca = prcomp(log2(datExpr+1), scale.=TRUE)

pca_plot = data.frame('ID'=rownames(datExpr), 'PC1'=datExpr_pca$x[,1], 'PC2'=datExpr_pca$x[,2], 
                      'meanExpr'=rowMeans(log2(datExpr+1)), 'sd' = apply(log2(datExpr+1), 1, sd)) %>%
           left_join(SFARI_genes, by='ID') %>%
           mutate(`gene-score` = ifelse(is.na(`gene-score`), 
                                        ifelse(ID %in% GO_neuronal$ID, 'Neuronal', 'Non-Neuronal'), 
                                        `gene-score`)) %>%
           dplyr::select(ID, PC1, PC2, `gene-score`, meanExpr, sd)

ggplotly(ggplot(pca_plot, aes(PC1, PC2, color=`gene-score`)) + geom_point(alpha=0.2, aes(id=ID)) + theme_minimal() +
         xlab(paste0('PC1 (',round(summary(datExpr_pca)$importance[2,1]*100),'%)')) +
         ylab(paste0('PC2 (',round(summary(datExpr_pca)$importance[2,2]*100),'%)')) +
         scale_color_manual(values=gg_colour_hue(9)))
```

91% of the variance of the probes seems to be explained by their mean level of expression
```{r}
ggplotly(ggplot(pca_plot, aes(PC1, PC2, color=meanExpr)) + geom_point(alpha=0.5, aes(id=ID)) + theme_minimal() +
         xlab(paste0('PC1 (',round(summary(datExpr_pca)$importance[2,1]*100),'%)')) +
         ylab(paste0('PC2 (',round(summary(datExpr_pca)$importance[2,2]*100),'%)')) +
         scale_colour_viridis())
```

The second PC seems to reflect the variance of the probe
```{r}
ggplotly(ggplot(pca_plot, aes(PC1, PC2, color=sd)) + geom_point(alpha=0.5, aes(id=ID)) + theme_minimal() +
         xlab(paste0('PC1 (',round(summary(datExpr_pca)$importance[2,1]*100),'%)')) +
         ylab(paste0('PC2 (',round(summary(datExpr_pca)$importance[2,2]*100),'%)')) +
         scale_colour_viridis())
```

#### Mean and SD by SFARI score

```{r, fig.width=10, fig.height=6, warning=FALSE}

ggplotly(pca_plot %>% filter(meanExpr<summary(pca_plot$meanExpr)[5]) %>% 
         ggplot(aes(meanExpr, color=`gene-score`, fill=`gene-score`)) +
         geom_histogram(binwidth=0.001, alpha=0.5) + facet_grid(`gene-score`~., scales='free') +
         theme_minimal() + ggtitle('mean expression for values up to 3rd quantile') +
         scale_color_manual(values=gg_colour_hue(9)) + scale_fill_manual(values=gg_colour_hue(9)))

```

Even after log2 transformation the variance is bigger for higher SFARI scores
```{r, fig.width=10, warning=FALSE}

make_Frontal_vs_Temporal_df = function(datExpr){
  datExpr_Frontal = datExpr %>% data.frame %>% dplyr::select(which(datMeta$Brain_lobe=='Frontal'))
  datExpr_Temporal = datExpr %>% data.frame %>% dplyr::select(which(datMeta$Brain_lobe=='Temporal'))
  
  Frontal_vs_Temporal = data.frame('ID'=as.character(rownames(datExpr)),
                          'mean' = rowMeans(datExpr), 'sd' = apply(datExpr, 1, sd),
                          'mean_Frontal'=rowMeans(datExpr_Frontal), 'mean_Temporal'=rowMeans(datExpr_Temporal),
                          'sd_Frontal'=apply(datExpr_Frontal,1,sd), 'sd_Temporal'=apply(datExpr_Temporal,1,sd)) %>%
               mutate('mean_diff'=mean_Frontal-mean_Temporal, 'sd_diff'=sd_Frontal-sd_Temporal) %>%
               left_join(SFARI_genes, by='ID') %>%
               dplyr::select(ID, mean, sd, mean_Frontal, mean_Temporal, mean_diff, sd_Frontal, sd_Temporal, sd_diff, `gene-score`) %>%
               mutate('Neuronal'=ifelse(ID %in% GO_neuronal$ID, 'Neuronal', 'Non-Neuronal'))
  
  Frontal_vs_Temporal_SFARI = Frontal_vs_Temporal %>% filter(!is.na(`gene-score`)) %>% 
                     mutate(Group = as.character(`gene-score`)) %>% dplyr::select(-c(Neuronal,`gene-score`))
  Frontal_vs_Temporal_Neuro = Frontal_vs_Temporal %>% mutate(Group = Neuronal) %>% dplyr::select(-c(Neuronal,`gene-score`))
  Frontal_vs_Temporal_all = Frontal_vs_Temporal %>% mutate(Group = 'All') %>% dplyr::select(-c(Neuronal,`gene-score`))
  
  Frontal_vs_Temporal_together = bind_rows(Frontal_vs_Temporal_SFARI, Frontal_vs_Temporal_Neuro, Frontal_vs_Temporal_all) %>% 
                        mutate(Group = ordered(Group, levels=c('1','2','3','4','5','6',
                                                               'Neuronal','Non-Neuronal','All')))
  
  return(Frontal_vs_Temporal_together)
}

Frontal_vs_Temporal = make_Frontal_vs_Temporal_df(log2(datExpr+1))

p1 = ggplotly(ggplot(Frontal_vs_Temporal, aes(Group, mean, fill=Group)) + 
              geom_boxplot() + theme_minimal() + 
              ggtitle('Mean (left) and Standard Deviation (right) by score') +
              scale_fill_manual(values=gg_colour_hue(9)) +
              theme(legend.position='none', axis.text.x=element_text(angle=45, hjust=1)))

p2 = ggplotly(ggplot(Frontal_vs_Temporal, aes(Group, sd, fill=Group)) + 
              geom_boxplot() + theme_minimal() +
              ggtitle('Mean (left) and Standard Deviation (right) by score') +
              scale_fill_manual(values=gg_colour_hue(9)) +
              theme(legend.position='none', axis.text.x=element_text(angle=45, hjust=1)))

subplot(p1, p2, nrows=1)

rm(p1, p2)
```

- Correlation and linear model fit only using SFARI and Neuronal genes

- The Mean increases 10x faster than the sd

- There's a weird pattern for the lowest expressed genes. Maybe their sd was topped to mean+n? Because the slope seems to be 1

```{r, warning=FALSE, fig.width=10}

fit = lm(sd ~ mean, data=Frontal_vs_Temporal[!Frontal_vs_Temporal$Group %in% c('Non-Neuronal','All'),])

ggplotly(ggplot(Frontal_vs_Temporal %>% filter(Group!='All'), aes(mean, sd)) + 
         geom_point(alpha=0.2, aes(id=ID, fill=Group, color=Group)) +
         scale_color_manual(values=gg_colour_hue(9)) + scale_fill_manual(values=gg_colour_hue(9)) +  
         scale_y_continuous(trans='log2', labels=scales::number_format(accuracy = 0.001)) + 
         scale_x_continuous(trans='log2', labels = scales::number_format(accuracy = 0.001)) + 
         ggtitle(paste0('Corr=',round(cor(Frontal_vs_Temporal$mean[!Frontal_vs_Temporal$Group %in% c('Non-Neuronal','All')], 
                                          Frontal_vs_Temporal$sd[!Frontal_vs_Temporal$Group %in% c('Non-Neuronal','All')]), 2),
                      ', LM fit: m=', round(fit$coefficients[[2]],4), ' b=', round(fit$coefficients[[1]],2))) +
         geom_abline(color='#808080', size=0.5) + theme_minimal())

```

## Filtering probes with low expression

Threshold: mean expression < 0.005 (8243 probes filtered out)

```{r}
to_keep = rowMeans(log2(datExpr+1))>0.005
datExpr = datExpr[to_keep,]
datProbes = datProbes[to_keep,]

rm(to_keep)
```

Number of genes and samples:
```{r}
dim(datExpr)
```
<br>

Gene count by SFARI score of remaining genes:
```{r}
table(SFARI_genes$`gene-score`[SFARI_genes$ID %in% rownames(datExpr)])
```
<br><br>

### Visualisations

#### PCA

Has exactly the same behaviour as before so didn't include the plots
<br>

#### Mean and SD by SFARI score

```{r, fig.width=10, warning=FALSE}
Frontal_vs_Temporal = make_Frontal_vs_Temporal_df(log2(datExpr+1))

p1 = ggplotly(ggplot(Frontal_vs_Temporal, aes(Group, mean, fill=Group)) + 
              geom_boxplot() + theme_minimal() + 
              ggtitle('Mean (left) and Standard Deviation (right) by score') +
              scale_fill_manual(values=gg_colour_hue(9)) +
              theme(legend.position='none', axis.text.x=element_text(angle=45, hjust=1)))

p2 = ggplotly(ggplot(Frontal_vs_Temporal, aes(Group, sd, fill=Group)) + 
              geom_boxplot() + theme_minimal() +
              ggtitle('Mean (left) and Standard Deviation (right) by score') +
              scale_fill_manual(values=gg_colour_hue(9)) +
              theme(legend.position='none', axis.text.x=element_text(angle=45, hjust=1)))

subplot(p1, p2, nrows=1)

rm(p1, p2)
```

Correlation and lm parameters almost don't change because they were calculated only using SFARI and Neuronal genes

```{r, warning=FALSE, fig.width=10}

fit = lm(sd ~ mean, data=Frontal_vs_Temporal[!Frontal_vs_Temporal$Group %in% c('Non-Neuronal','All'),])

ggplotly(ggplot(Frontal_vs_Temporal %>% filter(Group!='All'), aes(mean, sd)) + 
         geom_point(alpha=0.2, aes(id=ID, fill=Group, color=Group)) +
         scale_color_manual(values=gg_colour_hue(9)) + scale_fill_manual(values=gg_colour_hue(9)) +  
         scale_y_continuous(trans='log2', labels=scales::number_format(accuracy = 0.01)) + 
         scale_x_continuous(trans='log2', labels = scales::number_format(accuracy = 0.01)) + 
         ggtitle(paste0('Corr=',round(cor(Frontal_vs_Temporal$mean[!Frontal_vs_Temporal$Group %in% c('Non-Neuronal','All')], 
                                          Frontal_vs_Temporal$sd[!Frontal_vs_Temporal$Group %in% c('Non-Neuronal','All')]), 2),
                      ', LM fit: m=', round(fit$coefficients[[2]],4), ' b=', round(fit$coefficients[[1]],2))) +
         geom_abline(color='#808080', size=0.5) + theme_minimal())

```

## DE between brain lobes

```{r}

mod = model.matrix(~Brain_lobe, data=datMeta)
corfit = duplicateCorrelation(datExpr, mod, block=datMeta$donor_id)
lmfit = lmFit(datExpr, design=mod, block=datMeta$donor_id, correlation=corfit$consensus)

fit = eBayes(lmfit, trend=T, robust=T)
top_genes = topTable(fit, coef=2, number=nrow(datExpr))
DE_info = top_genes[match(rownames(datExpr), rownames(top_genes)),] %>% rownames_to_column('ID')

if(!file.exists('./../Data/BrainSpan_filtered.RData')){
  save(datExpr, datMeta, datProbes, SFARI_genes, GO_neuronal, file='./../Data/BrainSpan_filtered.RData')
}

rm(mod, corfit, lmfit, fit, top_genes)
```

The difference in mean is smaller the higher the SFARI score but the logFC is 
```{r, warning=FALSE, fig.width=10}
Frontal_vs_Temporal = make_Frontal_vs_Temporal_df(datExpr) %>% left_join(DE_info, by='ID')

p1 = ggplotly(ggplot(Frontal_vs_Temporal, aes(Group, abs(mean_diff), fill=Group)) + 
                geom_boxplot() + theme_minimal() + 
                ggtitle('Difference in mean (left) and log Fold Change (right) by Brain region') +
                scale_fill_manual(values=gg_colour_hue(9)) +
                theme(legend.position='none', axis.text.x=element_text(angle=45, hjust=1)))

p2 = ggplotly(ggplot(Frontal_vs_Temporal, aes(Group, abs(logFC), fill=Group)) + 
                geom_boxplot() + theme_minimal() +
                ggtitle('Difference in mean (left) and log Fold Change (right) by Brain region') +
                scale_fill_manual(values=gg_colour_hue(9)) +
                theme(legend.position='none', axis.text.x=element_text(angle=45, hjust=1)))

subplot(p1, p2, nrows=1)

rm(p1, p2)
```







<!-- ### Boxplots for original data -->

<!-- #### Mean and Standard Deviation by score -->

<!-- - The higher the SFARI score, the higher the mean and the standard deviation except for the sd of SFARI scores 4 and 5 -->
<!-- ```{r, fig.width=10, warning=FALSE} -->

<!-- make_Frontal_vs_Temporal_df = function(datExpr){ -->
<!--   datExpr_Frontal = datExpr %>% data.frame %>% dplyr::select(which(datMeta$Brain_lobe=='Frontal')) -->
<!--   datExpr_Temporal = datExpr %>% data.frame %>% dplyr::select(which(datMeta$Brain_lobe=='Temporal')) -->

<!--   Frontal_vs_Temporal = data.frame('ID'=as.character(rownames(datExpr)), -->
<!--                           'mean' = rowMeans(datExpr), 'sd' = apply(datExpr, 1, sd), -->
<!--                           'mean_Frontal'=rowMeans(datExpr_Frontal), 'mean_Temporal'=rowMeans(datExpr_Temporal), -->
<!--                           'sd_Frontal'=apply(datExpr_Frontal,1,sd), 'sd_Temporal'=apply(datExpr_Temporal,1,sd)) %>% -->
<!--                mutate('mean_diff'=mean_Frontal-mean_Temporal, 'sd_diff'=sd_Frontal-sd_Temporal) %>% -->
<!--                left_join(SFARI_genes, by='ID') %>% -->
<!--                dplyr::select(ID, mean, sd, mean_Frontal, mean_Temporal, mean_diff, sd_Frontal, sd_Temporal, sd_diff, `gene-score`) %>% -->
<!--                mutate('Neuronal'=ifelse(ID %in% GO_neuronal$ID, 'Neuronal', 'Non-Neuronal')) -->

<!--   Frontal_vs_Temporal_SFARI = Frontal_vs_Temporal %>% filter(!is.na(`gene-score`)) %>%  -->
<!--                      mutate(Group = as.character(`gene-score`)) %>% dplyr::select(-c(Neuronal,`gene-score`)) -->
<!--   Frontal_vs_Temporal_Neuro = Frontal_vs_Temporal %>% mutate(Group = Neuronal) %>% dplyr::select(-c(Neuronal,`gene-score`)) -->
<!--   Frontal_vs_Temporal_all = Frontal_vs_Temporal %>% mutate(Group = 'All') %>% dplyr::select(-c(Neuronal,`gene-score`)) -->

<!--   Frontal_vs_Temporal_together = bind_rows(Frontal_vs_Temporal_SFARI, Frontal_vs_Temporal_Neuro, Frontal_vs_Temporal_all) %>%  -->
<!--                         mutate(Group = ordered(Group, levels=c('1','2','3','4','5','6', -->
<!--                                                                'Neuronal','Non-Neuronal','All'))) %>% -->
<!--                         left_join(DE_info, by='ID') -->

<!--   return(Frontal_vs_Temporal_together) -->
<!-- } -->

<!-- Frontal_vs_Temporal = make_Frontal_vs_Temporal_df(datExpr) -->

<!-- p1 = ggplotly(ggplot(Frontal_vs_Temporal, aes(Group, mean, fill=Group)) +  -->
<!--               geom_boxplot() + theme_minimal() +  -->
<!--               ggtitle('Mean (left) and Standard Deviation (right) by score') + -->
<!--               scale_fill_manual(values=gg_colour_hue(9)) + -->
<!--               theme(legend.position='none', axis.text.x=element_text(angle=45, hjust=1))) -->

<!-- p2 = ggplotly(ggplot(Frontal_vs_Temporal, aes(Group, sd, fill=Group)) +  -->
<!--               geom_boxplot() + theme_minimal() + -->
<!--               ggtitle('Mean (left) and Standard Deviation (right) by score') + -->
<!--               scale_fill_manual(values=gg_colour_hue(9)) + -->
<!--               theme(legend.position='none', axis.text.x=element_text(angle=45, hjust=1))) -->

<!-- subplot(p1, p2, nrows=1) -->

<!-- rm(p1, p2) -->
<!-- ``` -->


<!-- ```{r, warning=FALSE, fig.width=10} -->
<!-- ggplotly(ggplot(Frontal_vs_Temporal %>% filter(Group!='All'), aes(mean, sd)) + geom_point(alpha=0.2, aes(id=ID, fill=Group, color=Group)) + -->
<!--          scale_color_manual(values=gg_colour_hue(9)) + scale_fill_manual(values=gg_colour_hue(9)) +   -->
<!--          scale_y_continuous(trans='log2') + #scale_x_continuous(trans='log2') +  -->
<!--            ggtitle(paste0('Corr=',round(cor(Frontal_vs_Temporal$mean, Frontal_vs_Temporal$sd), 2))) + -->
<!--          geom_abline(color='#808080', size=0.5) + theme_minimal()) -->

<!-- lm(sd~mean, data=Frontal_vs_Temporal)$coefficients -->
<!-- ``` -->

<!-- #### Difference between Brain lobe by score -->

<!-- ```{r, fig.width=10, warning=FALSE} -->
<!-- p1 = ggplotly(ggplot(Frontal_vs_Temporal, aes(Group, abs(mean_diff), fill=Group)) +  -->
<!--                 geom_boxplot() + theme_minimal() +  -->
<!--                 ggtitle('Difference in mean (left) and log Fold Change (right) by Brain region') + -->
<!--                 scale_fill_manual(values=gg_colour_hue(9)) + -->
<!--                 theme(legend.position='none', axis.text.x=element_text(angle=45, hjust=1))) -->

<!-- p2 = ggplotly(ggplot(Frontal_vs_Temporal, aes(Group, abs(logFC), fill=Group)) +  -->
<!--                 geom_boxplot() + theme_minimal() + -->
<!--                 ggtitle('Difference in mean (left) and log Fold Change (right) by Brain region') + -->
<!--                 scale_fill_manual(values=gg_colour_hue(9)) + -->
<!--                 theme(legend.position='none', axis.text.x=element_text(angle=45, hjust=1))) -->

<!-- subplot(p1, p2, nrows=1) -->

<!-- rm(p1, p2) -->
<!-- ``` -->

