---
title: 'DESeq2 vs limma'
output:
  html_document:
    code_folding: 'hide'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=TRUE, include=FALSE}
setwd('/afs/inf.ed.ac.uk/user/s17/s1725186/Documents/PhD-InitialExperiments/Gandal/RNAseq/R_markdowns')
#setwd('/home/magdalena/PhD/initialExperiments/Gandal/RNAseq/R_markdowns')

library(limma); library(cqn) ; library(DESeq2)
library(tidyverse) ; library(reshape2) ; library(glue) ; library(plotly) ; library(dendextend)
library(RColorBrewer) ; library(viridis) ; require(gridExtra) ; 
library(plotlyutils) # https://github.com/Alanocallaghan/plotlyutils/
library(biomaRt)

gg_colour_hue = function(n) {
  hues = seq(15, 375, length = n-1)
  pal = c(hcl(h = hues, l = 65, c = 100)[1:(n-2)], 'gray','#808080')
}
```

```{r, message=FALSE}

# Load csvs
datExpr = read.csv('./../raw_data/RNAseq_ASD_datExpr.csv', row.names=1)
datMeta = read.csv('./../raw_data/RNAseq_ASD_datMeta.csv')
SFARI_genes = read_csv('./../working_data/SFARI_genes_01-15-2019.csv')

# Make sure datExpr and datMeta columns/rows match
rownames(datMeta) = paste0('X', datMeta$Dissected_Sample_ID)
if(!all(colnames(datExpr) == rownames(datMeta))){
  print('Columns in datExpr don\'t match the rowd in datMeta!')
}

# Annotate probes
getinfo = c('ensembl_gene_id','external_gene_id','chromosome_name','start_position',
            'end_position','strand','band','gene_biotype','percentage_gc_content')
mart = useMart(biomart='ENSEMBL_MART_ENSEMBL',
               dataset='hsapiens_gene_ensembl',
               host='feb2014.archive.ensembl.org') ## Gencode v19
datProbes = getBM(attributes=getinfo, filters=c('ensembl_gene_id'), values=rownames(datExpr), mart=mart)
datProbes = datProbes[match(rownames(datExpr), datProbes$ensembl_gene_id),]
datProbes$length = datProbes$end_position-datProbes$start_position

# Group brain regions by lobes
datMeta$Brain_Region = as.factor(datMeta$Region)
datMeta$Brain_lobe = 'Occipital'
datMeta$Brain_lobe[datMeta$Brain_Region %in% c('BA4_6', 'BA9', 'BA24', 'BA44_45')] = 'Frontal'
datMeta$Brain_lobe[datMeta$Brain_Region %in% c('BA3_1_2_5', 'BA7')] = 'Parietal'
datMeta$Brain_lobe[datMeta$Brain_Region %in% c('BA38', 'BA39_40', 'BA20_37', 'BA41_42_22')] = 'Temporal'
datMeta$Brain_lobe=factor(datMeta$Brain_lobe, levels=c('Frontal', 'Temporal', 'Parietal', 'Occipital'))

#################################################################################
# FILTERS:

# 1 Filter probes with start or end position missing (filter 5)
# These can be filtered without probe info, they have weird IDS that don't start with ENS
to_keep = !is.na(datProbes$length)
datProbes = datProbes[to_keep,]
datExpr = datExpr[to_keep,]
rownames(datProbes) = datProbes$ensembl_gene_id

# 2. Filter samples from ID AN03345 (filter 2)
to_keep = (datMeta$Subject_ID != 'AN03345')
datMeta = datMeta[to_keep,]
datExpr = datExpr[,to_keep]

# 3. Filter samples with rowSums <= 40
to_keep = rowSums(datExpr)>40
datExpr = datExpr[to_keep,]
datProbes = datProbes[to_keep,]

#################################################################################
# Annotate SFARI genes

# Get ensemble IDS for SFARI genes
mart = useMart(biomart='ENSEMBL_MART_ENSEMBL',
               dataset='hsapiens_gene_ensembl',
               host='feb2014.archive.ensembl.org') ## Gencode v19

gene_names = getBM(attributes=c('ensembl_gene_id', 'hgnc_symbol'), filters=c('hgnc_symbol'), 
                   values=SFARI_genes$`gene-symbol`, mart=mart) %>% 
                   mutate('gene-symbol'=hgnc_symbol, 'ID'=as.character(ensembl_gene_id)) %>% 
                   dplyr::select('ID', 'gene-symbol')

SFARI_genes = left_join(SFARI_genes, gene_names, by='gene-symbol')

#################################################################################
# Add functional annotation to genes from GO

GO_annotations = read.csv('./../working_data/genes_GO_annotations.csv')
GO_neuronal = GO_annotations %>% filter(grepl('neuron', go_term)) %>% 
              mutate('ID' = as.character(ensembl_gene_id)) %>% 
              dplyr::select(-ensembl_gene_id) %>% distinct(ID) %>%
              mutate('Neuronal' = 1)


datExpr_backup = datExpr

rm(getinfo, to_keep, gene_names, mart, GO_annotations)

```

```{r}
################################################################################
# Calculate Differential Expression using limma and DESeq2 packages

# Limma
if(!file.exists('./../working_data/genes_ASD_DE_info_limma.csv')){
  
  # First perform cqn normalisation
  cqn.dat = cqn(datExpr, lengths = as.numeric(datProbes$length),
                x = as.numeric(datProbes$percentage_gc_content),
                lengthMethod = c('smooth'),
                sqn = FALSE)
  datExpr_cqn = cqn.dat$y + cqn.dat$offset
  
  mod = model.matrix(~ Diagnosis_, data=datMeta)
  corfit = duplicateCorrelation(datExpr_cqn, mod, block=datMeta$Subject_ID)
  lmfit = lmFit(datExpr_cqn, design=mod, block=datMeta$Subject_ID, correlation=corfit$consensus)
  fit = eBayes(lmfit, trend=T, robust=T)
  top_genes = topTable(fit, coef=2, number=nrow(datExpr_cqn))
  DE_info_limma = top_genes[match(rownames(datExpr_cqn), rownames(top_genes)),] %>%
                  rownames_to_column(var = 'ID') %>%
                  mutate('logFC_limma'=logFC, 'adj.P.Val_limma'=adj.P.Val) %>%
                  dplyr::select(ID, logFC_limma, adj.P.Val_limma)  
  
  write.csv(DE_info_limma, './../working_data/genes_ASD_DE_info_limma.csv')
  
  rm(mod, corfit, lmfit, fit, top_genes)
  
} else DE_info_limma = read.csv('./../working_data/genes_ASD_DE_info_limma.csv')

# DESeq2 DE
if(!file.exists('./../working_data/genes_ASD_DE_info_DESeq2.csv')){
  counts = as.matrix(datExpr)
  rowRanges = GRanges(datProbes$chromosome_name,
                      IRanges(datProbes$start_position, width=datProbes$length),
                      strand=datProbes$strand,
                      feature_id=datProbes$ensembl_gene_id)
  
  se = SummarizedExperiment(assays=SimpleList(counts=counts), rowRanges=rowRanges, colData=datMeta)
  ddsSE = DESeqDataSet(se, design =~Diagnosis_)
  
  dds = DESeq(ddsSE)
  DE_info_DESeq2 = results(dds) %>% data.frame %>% rownames_to_column(var = 'ID') %>%
                   mutate('logFC_DESeq2'=log2FoldChange, 'adj.P.Val_DESeq2'=padj) %>% 
                   dplyr::select(ID, logFC_DESeq2, adj.P.Val_DESeq2)
  
  write.csv(DE_info_DESeq2, './../working_data/genes_ASD_DE_info_DESeq2.csv')
  
  rm(counts, rowRanges, se, ddsSE, dds)
  
} else DE_info_DESeq2 = read.csv('./../working_data/genes_ASD_DE_info_DESeq2.csv')

DE_info = DE_info_limma %>% left_join(DE_info_DESeq2, by='ID')

rm(DE_info_limma, DE_info_DESeq2)

```
<br>

## DE Output Comparisons

#### **Log fold change**
```{r, warning=FALSE}
ggplotly(DE_info %>% ggplot(aes(logFC_DESeq2, logFC_limma)) + geom_point(alpha=0.05, aes(id=ID)) +
         geom_vline(xintercept=-log2(1.2), color='#cc0099', size=0.5) + 
         geom_hline(yintercept=-log2(1.2), color='#cc0099', size=0.5) +
         geom_vline(xintercept= log2(1.2), color='#cc0099', size=0.5) + 
         geom_hline(yintercept= log2(1.2), color='#cc0099', size=0.5) +
         geom_smooth(method=lm, se=F, color='#009999', size=0.5) + theme_minimal() + coord_fixed() +
         ggtitle(glue('LogFC (corr=',round(cor(DE_info$logFC_limma, DE_info$logFC_DESeq2),2),')')))
```

#### **P-values**

Although both functions use the BH method to adjust for multiple testing, the limma package seems to be more strict, assigning a high p-value to many of the points that DESeq2 considers statistically significant
```{r, warning=FALSE}
DE_info %>% ggplot(aes(adj.P.Val_DESeq2, adj.P.Val_limma)) + geom_point(alpha=0.05, aes(id=ID)) + 
            geom_vline(xintercept=0.05, color='#cc0099') + geom_hline(yintercept=0.05, color='#cc0099') +
            geom_smooth(method=lm, se=F, color='#009999', size=0.5) + theme_minimal() + coord_fixed() +
            ggtitle(glue('Adjusted p-values (corr=',round(cor(DE_info$adj.P.Val_limma, 
                                                                DE_info$adj.P.Val_DESeq2),2),')'))
```

##### Statistical significance concordance using adjusted p-values:
```{r}
p_val_significance = data.frame('limma Significance' = DE_info$adj.P.Val_limma<0.05,
                                'DESeq2 Significance' = DE_info$adj.P.Val_DESeq2<0.05)
table(p_val_significance)

print(glue(round(sum(DE_info$adj.P.Val_limma<0.05 & DE_info$adj.P.Val_DESeq2>0.05)/sum(DE_info$adj.P.Val_limma<0.05)*100,2),
           '% of the statistically significant limma genes are not significant for DESeq2.
           ', round(sum(DE_info$adj.P.Val_limma>0.05 & DE_info$adj.P.Val_DESeq2<0.05)/sum(DE_info$adj.P.Val_DESeq2<0.05)*100,2),
           '% of the statistically significant DESeq2 genes are not significant for limma.'))

rm(p_val_significance)
```
<br><br>

## Transform expression levels from ASD to match CTL

Calculating the mean expression of the genes that don't have a neuronal functional annotation nor a SFARI score for ADS and CTL separately and plotting ASD vs CTL with log2 scale on both axis we can see that the fitted line has a slope close to 0.5
```{r, warning=FALSE}
housekeeping_genes = rownames(datExpr)[!rownames(datExpr) %in% GO_neuronal$ID & 
                                       !rownames(datExpr) %in% SFARI_genes$ID]
housekeeping_genes_ASD = datExpr %>% dplyr::select(which(datMeta$Diagnosis_=='ASD')) %>% 
                              filter(rownames(.) %in% housekeeping_genes) %>% rowMeans
housekeeping_genes_CTL = datExpr %>% dplyr::select(which(datMeta$Diagnosis_!='ASD')) %>% 
                              filter(rownames(.) %in% housekeeping_genes) %>% rowMeans

ASD_vs_CTL = data.frame('ID'=housekeeping_genes, 'ASD'=housekeeping_genes_ASD+1, 'CTL'=housekeeping_genes_CTL+1)

ggplotly(ASD_vs_CTL %>% ggplot(aes(ASD, CTL)) + geom_point(alpha=0.1, aes(id=ID)) + 
               geom_smooth(method=lm, se=F, color='#009999') + theme_minimal() +
               scale_x_continuous(trans='log2') + scale_y_continuous(trans='log2'))


rm(housekeeping_genes_ASD, housekeeping_genes_CTL)
```

#### Transformation:

Fit a linear model between ASD and CTL for these non-neuronal, non-SFARI genes and use the slope and intercept to transform the ASD expression to match the one from the CTL samples.
```{r}
fit = lm(log2(ASD) ~ log2(CTL), ASD_vs_CTL)
slope = summary(fit)$coefficients[2,1]
intercept = summary(fit)$coefficients[1,1]

# print(glue('Intercept = ', round(intercept,2), ' and slope = ', round(slope,2)))

datExpr_ASD = datExpr %>% dplyr::select(which(datMeta$Diagnosis_=='ASD'))
datExpr_CTL = datExpr %>% dplyr::select(which(datMeta$Diagnosis_!='ASD'))

datExpr_ASD = (datExpr_ASD-intercept)/slope

datExpr_fix = bind_cols(datExpr_ASD, datExpr_CTL)
datExpr_fix = datExpr_fix[,match(colnames(datExpr),colnames(datExpr_fix))]
rownames(datExpr_fix) = rownames(datExpr)

housekeeping_genes_ASD = datExpr_fix %>% dplyr::select(which(datMeta$Diagnosis_=='ASD')) %>% 
                              filter(rownames(.) %in% housekeeping_genes) %>% rowMeans
housekeeping_genes_CTL = datExpr_fix %>% dplyr::select(which(datMeta$Diagnosis_!='ASD')) %>% 
                              filter(rownames(.) %in% housekeeping_genes) %>% rowMeans

ASD_vs_CTL = data.frame('ASD' = housekeeping_genes_ASD+1, 'CTL' = housekeeping_genes_CTL+1)

ggplotly(ASD_vs_CTL %>% sample_n(10000) %>% ggplot(aes(ASD, CTL)) + geom_point(alpha=0.1) + 
         geom_smooth(method=lm, se=F, color='#009999') + scale_x_continuous(trans='log2') + 
         scale_y_continuous(trans='log2') + theme_minimal())

rm(fit, slope, intercept, datExpr_ASD, datExpr_CTL, housekeeping_genes, 
   housekeeping_genes_ASD, housekeeping_genes_CTL, ASD_vs_CTL)
```

### Differential Expression analysis with transformed data

* We need to round the numbers to treat them as raw counts for the DESeq2 package and make them all > 0

```{r}

min_val = datExpr_fix %>% apply(1, min) %>% min
if(min_val<0) datExpr_fix = datExpr_fix + abs(min_val)
datExpr_fix = round(datExpr_fix)

datExpr = datExpr_fix

rm(min_val)
```

```{r, warning=FALSE}
################################################################################
# Calculate Differential Expression using limma and DESeq2 packages

# Limma
if(!file.exists('./../working_data/genes_ASD_DE_info_limma_fix.csv')){
  
  # First perform cqn normalisation
  cqn.dat = cqn(datExpr, lengths = as.numeric(datProbes$length),
                x = as.numeric(datProbes$percentage_gc_content),
                lengthMethod = c('smooth'),
                sqn = FALSE)
  datExpr_cqn = cqn.dat$y + cqn.dat$offset
  
  mod = model.matrix(~ Diagnosis_, data=datMeta)
  corfit = duplicateCorrelation(datExpr_cqn, mod, block=datMeta$Subject_ID)
  lmfit = lmFit(datExpr_cqn, design=mod, block=datMeta$Subject_ID, correlation=corfit$consensus)
  fit = eBayes(lmfit, trend=T, robust=T)
  top_genes = topTable(fit, coef=2, number=nrow(datExpr_cqn))
  DE_info_limma = top_genes[match(rownames(datExpr_cqn), rownames(top_genes)),] %>%
                  rownames_to_column(var = 'ID') %>%
                  mutate('logFC_limma'=logFC, 'adj.P.Val_limma'=adj.P.Val) %>%
                  dplyr::select(ID, logFC_limma, adj.P.Val_limma) 
  
  write.csv(DE_info_limma, './../working_data/genes_ASD_DE_info_limma_fix.csv')
  
  rm(mod, corfit, lmfit, fit, top_genes)
  
} else DE_info_limma = read.csv('./../working_data/genes_ASD_DE_info_limma_fix.csv')

# DESeq2 DE
if(!file.exists('./../working_data/genes_ASD_DE_info_DESeq2_fix.csv')){
  counts = as.matrix(datExpr)
  rowRanges = GRanges(datProbes$chromosome_name,
                      IRanges(datProbes$start_position, width=datProbes$length),
                      strand=datProbes$strand,
                      feature_id=datProbes$ensembl_gene_id)
  
  se = SummarizedExperiment(assays=SimpleList(counts=counts), rowRanges=rowRanges, colData=datMeta)
  ddsSE = DESeqDataSet(se, design =~Diagnosis_)
  
  dds = DESeq(ddsSE)
  DE_info_DESeq2 = results(dds) %>% data.frame %>% rownames_to_column(var = 'ID') %>%
                   mutate('logFC_DESeq2'=log2FoldChange, 'adj.P.Val_DESeq2'=padj) %>% 
                   dplyr::select(ID, logFC_DESeq2, adj.P.Val_DESeq2)
  
  write.csv(DE_info_DESeq2, './../working_data/genes_ASD_DE_info_DESeq2_fix.csv')
  
  rm(counts, rowRanges, se, ddsSE, dds)
  
} else DE_info_DESeq2 = read.csv('./../working_data/genes_ASD_DE_info_DESeq2_fix.csv')

DE_info = DE_info_limma %>% left_join(DE_info_DESeq2, by='ID')

rm(DE_info_limma, DE_info_DESeq2)

```
<br>

#### DE Output Comparisons

##### **Log fold change**
```{r, warning=FALSE}
ggplotly(DE_info %>% ggplot(aes(logFC_DESeq2, logFC_limma)) + geom_point(alpha=0.05, aes(id=ID)) +
         geom_vline(xintercept=-log2(1.2), color='#cc0099', size=0.5) + 
         geom_hline(yintercept=-log2(1.2), color='#cc0099', size=0.5) +
         geom_vline(xintercept= log2(1.2), color='#cc0099', size=0.5) + 
         geom_hline(yintercept= log2(1.2), color='#cc0099', size=0.5) +
         ggtitle(glue('LogFC (corr=',round(cor(DE_info$logFC_limma, DE_info$logFC_DESeq2),2),')')) +
         geom_smooth(method=lm, se=F, color='#009999', size=0.5) + coord_fixed() +
         theme_minimal())
```

##### **P-values**

Although both functions use the BH method to adjust for multiple testing, the limma package seems to be more strict, assigning a high p-value to many of the points that DESeq2 considers statistically significant

```{r, warning=FALSE}
DE_info %>% ggplot(aes(adj.P.Val_DESeq2, adj.P.Val_limma)) + geom_point(alpha=0.05, aes(id=ID)) + 
            geom_vline(xintercept=0.05, color='#cc0099') + 
            geom_hline(yintercept=0.05, color='#cc0099') + 
            geom_smooth(method=lm, se=F, color='#009999', size=0.5) + coord_fixed() +
            ggtitle(glue('Adjusted p-values (corr=',round(cor(DE_info$adj.P.Val_limma, 
                                                           DE_info$adj.P.Val_DESeq2),2),')')) + 
            theme_minimal()
```

##### Statistical significance concordance using adjusted p-values
```{r}
p_val_significance = data.frame('limma Significance' = DE_info$adj.P.Val_limma<0.05,
                                'DESeq2 Significance' = DE_info$adj.P.Val_DESeq2<0.05)
table(p_val_significance)

print(glue(round(sum(DE_info$adj.P.Val_limma<0.05 & DE_info$adj.P.Val_DESeq2>0.05)/sum(DE_info$adj.P.Val_limma<0.05)*100,2),
           '% of the statistically significant limma genes are not significant for DESeq2.
           ', round(sum(DE_info$adj.P.Val_limma>0.05 & DE_info$adj.P.Val_DESeq2<0.05)/sum(DE_info$adj.P.Val_DESeq2<0.05)*100,2),
           '% of the statistically significant DESeq2 genes are not significant for limma.'))

rm(p_val_significance)
```
<br><br>

## Compare within methods before and after transformation

### Limma
```{r, warning=FALSE}
limma_orig = read.csv('./../working_data/genes_ASD_DE_info_limma.csv') %>% dplyr::select(-X)
colnames(limma_orig) = c('ID','logFC_orig', 'adj.P.Val_orig')

limma_fix = read.csv('./../working_data/genes_ASD_DE_info_limma_fix.csv') %>% dplyr::select(-X)
colnames(limma_fix) = c('ID','logFC_fix', 'adj.P.Val_fix')

limma_both = limma_orig %>% left_join(limma_fix, by='ID')

ggplotly(limma_both %>% ggplot(aes(logFC_orig, logFC_fix)) + geom_point(alpha=0.05, aes(id=ID)) + coord_fixed() +
         geom_vline(xintercept=-log2(1.2), color='#cc0099', size=0.5) + 
         geom_hline(yintercept=-log2(1.2), color='#cc0099', size=0.5) +
         geom_vline(xintercept= log2(1.2), color='#cc0099', size=0.5) + 
         geom_hline(yintercept= log2(1.2), color='#cc0099', size=0.5) +
         theme_minimal()+  ggtitle(glue('LogFC (corr=',round(cor(limma_both$logFC_orig, 
                                                                 limma_both$logFC_fix),4),')')))

ggplotly(limma_both %>% ggplot(aes(adj.P.Val_orig, adj.P.Val_fix)) + geom_point(alpha=0.05, aes(id=ID)) +
         geom_vline(xintercept=0.05, color='#cc0099') + geom_hline(yintercept=0.05, color='#cc0099') +
         theme_minimal()+  ggtitle(glue('adjusted p-value (corr=',round(cor(limma_both$adj.P.Val_orig, 
                                        limma_both$adj.P.Val_fix),4),')')) + coord_fixed())

rm(limma_orig, limma_fix)
```

```{r}
limma_comp = data.frame('orig_significant' = limma_both$logFC_orig>log2(1.2) & limma_both$adj.P.Val_orig<0.05,
                        'fix_significant' = limma_both$logFC_fix>log2(1.2) & limma_both$adj.P.Val_fix<0.05)

table(limma_comp)
```

### DESeq2
```{r, warning=FALSE}
DESeq2_orig = read.csv('./../working_data/genes_ASD_DE_info_DESeq2.csv') %>% dplyr::select(-X)
colnames(DESeq2_orig) = c('ID','logFC_orig', 'adj.P.Val_orig')

DESeq2_fix = read.csv('./../working_data/genes_ASD_DE_info_DESeq2_fix.csv') %>% dplyr::select(-X)
colnames(DESeq2_fix) = c('ID','logFC_fix', 'adj.P.Val_fix')

DESeq2_both = DESeq2_orig %>% left_join(DESeq2_fix, by='ID')

ggplotly(DESeq2_both %>% ggplot(aes(logFC_orig, logFC_fix)) + geom_point(alpha=0.05, aes(id=ID)) + coord_fixed() +
         geom_vline(xintercept=-log2(1.2), color='#cc0099', size=0.5) + 
         geom_hline(yintercept=-log2(1.2), color='#cc0099', size=0.5) +
         geom_vline(xintercept= log2(1.2), color='#cc0099', size=0.5) + 
         geom_hline(yintercept= log2(1.2), color='#cc0099', size=0.5) +
         theme_minimal() + ggtitle(glue('LogFC (corr=',round(cor(DESeq2_both$logFC_orig, 
                                                                 DESeq2_both$logFC_fix),4),')')))

ggplotly(DESeq2_both %>% ggplot(aes(adj.P.Val_orig, adj.P.Val_fix)) + geom_point(alpha=0.05, aes(id=ID)) +
         geom_vline(xintercept=0.05, color='#cc0099') + geom_hline(yintercept=0.05, color='#cc0099') +
         theme_minimal()+  ggtitle(glue('adjusted p-value (corr=',round(cor(DESeq2_both$adj.P.Val_orig, 
                                        DESeq2_both$adj.P.Val_fix),4),')')) + coord_fixed())

rm(DESeq2_orig, DESeq2_fix)
```

```{r}
DESeq2_both = data.frame('orig_significant' = DESeq2_both$logFC_orig>log2(1.2) & DESeq2_both$adj.P.Val_orig<0.05,
                        'fix_significant' = DESeq2_both$logFC_fix>log2(1.2) & DESeq2_both$adj.P.Val_fix<0.05)

table(DESeq2_both)
```




