---
title: "Clustering Genes"
# author: "Magdalena Navarro"
date: "19 March 2019"
output:
  html_document:
    code_folding: "hide"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Filter criteria:

- Filter differentially expressed genes between autism and control (p-value < 0.05)

- No samples are removed based on network connectivity z-scores

```{r, echo=TRUE, include=FALSE}
setwd('/afs/inf.ed.ac.uk/user/s17/s1725186/Documents/PhD-InitialExperiments/Gandal/RNAseq/R_markdowns')
#setwd('/home/magdalena/PhD/initialExperiments/Gandal/RNAseq/R_markdowns')

library(limma);
library(tidyverse) ; library(reshape2) ; library(glue) ; library(plotly) ; library(dendextend)
library(RColorBrewer) ; library(viridis) ; require(gridExtra)
library(ConsensusClusterPlus)
library(JADE) ; library(MineICA) ; library(moments) ; library(fdrtool)
library(ClusterR)
library(WGCNA)
library(pdfCluster) ; library(gplots)

# # FILTER DE GENES
# load('./../working_data/RNAseq_ASD_4region_normalized.Rdata')
# 
# # Balance Groups by covariates, remove singular batches (none)
# to_keep = (datMeta$Subject_ID != 'AN03345') & !is.na(datMeta$Dx)
# table(to_keep)
# datMeta = datMeta[to_keep,]
# datExpr = datExpr[,to_keep]
# datSeq = datSeq[to_keep,]
# 
# # Select genes differentially expressed in ASD
# mod = model.matrix(~ Dx, data=datMeta)
# corfit = duplicateCorrelation(datExpr, mod, block=datMeta$Subject_ID)
# lmfit = lmFit(datExpr, design=mod, block=datMeta$Subject_ID, correlation=corfit$consensus)
# 
# fit = eBayes(lmfit, trend=T, robust=T)
# top_genes = topTable(fit, coef=2, number=nrow(datExpr))
# ordered_top_genes = top_genes[match(rownames(datExpr), rownames(top_genes)),]
# 
# ASD_pvals = rownames(ordered_top_genes)[ordered_top_genes$adj.P.Val<0.05]     # 17% of genes (3385)
# ASD_lfc = rownames(ordered_top_genes)[abs(ordered_top_genes$logFC)>log2(1.2)] # 33% of genes (7048)
# ASD_pvals_lfc = intersect(ASD_pvals, ASD_lfc)                                 # 14% of genes (2928)
# 
# datExpr = datExpr[match(ASD_pvals_lfc, rownames(datExpr)),]
#
# rm(corfit, fit, lmfit, mod, ASD_pvals, ASD_lfc, ASD_pvals_lfc, top_genes)

load('./../working_data/RNAseq_ASD_4region_DEgenes_adj_pval_lfc.Rdata')
```

```{r}
glue('Number of genes: ', nrow(datExpr), '\n',
     'Number of samples: ', ncol(datExpr), ' (', sum(datMeta$Diagnosis_=='ASD'), ' ASD, ',
     sum(datMeta$Diagnosis_!='ASD'), ' controls)')
```

### Dimensionality reduction using PCA

First principal component explains over 89.7% of the total variance (lower than 90%...)

```{r, fig.width = 12}
reduce_dim_datExpr = function(datExpr, datMeta, var_explained=0.98){
  
  datExpr = data.frame(datExpr)
  
  datExpr_pca = prcomp(datExpr, scale.=TRUE)
  last_pc = data.frame(summary(datExpr_pca)$importance[3,]) %>% rownames_to_column(var='ID') %>% 
    filter(.[[2]] >= var_explained) %>% top_n(-1, ID)
  
  par(mfrow=c(1,2))
  plot(summary(datExpr_pca)$importance[2,], type='b')
  abline(v=substr(last_pc$ID, 3, nchar(last_pc$ID)), col='blue')
  plot(summary(datExpr_pca)$importance[3,], type='b')
  abline(h=var_explained, col='blue')
  
  
  print(glue('Keeping top ', substr(last_pc$ID, 3, nchar(last_pc$ID)), ' components that explain ',
             var_explained*100, '% of the variance'))
  
  datExpr_top_pc = datExpr_pca$x %>% data.frame %>% dplyr::select(PC1:last_pc$ID)
  
  return(list('datExpr'=datExpr_top_pc, 'pca_output'=datExpr_pca))
}

reduce_dim_output = reduce_dim_datExpr(datExpr, datMeta)
datExpr_redDim = reduce_dim_output$datExpr
pca_output = reduce_dim_output$pca_output

rm(datSeq, datProbes, reduce_dim_datExpr, reduce_dim_output, datExpr)
```

Genes are still separated into two clouds of points:
```{r}
datExpr_redDim %>% ggplot(aes(x=PC1, y=PC2)) + geom_point() + theme_minimal()
```

Projecting all the original points into the space created by the two principal components and colouring by the differential expression p-value we can see that the points in the middle of the two clouds were filtered out because their DE wasn't statistically significant. Colouring by their log2 fold change we can see that the genes from the cloud on the top are overexpressed and the genes in the bottom one underexpressed.
```{r, fig.width = 12}
load('./../working_data/RNAseq_ASD_4region_normalized.Rdata')

# Remove Subject with ID = 'AN03345'
keep = datMeta$Subject_ID!='AN03345'
datMeta = datMeta[keep,]
datExpr = datExpr[,keep]

# Calculate differential expression p-value of each gene
mod = model.matrix(~ Dx, data=datMeta)
corfit = duplicateCorrelation(datExpr, mod, block=datMeta$Subject_ID)
lmfit = lmFit(datExpr, design=mod, block=datMeta$Subject_ID, correlation=corfit$consensus)

fit = eBayes(lmfit, trend=T, robust=T)
top_genes = topTable(fit, coef=2, number=nrow(datExpr))
ordered_top_genes = top_genes[match(rownames(datExpr), rownames(top_genes)),]
  
ASD_pvals = ordered_top_genes$adj.P.Val
log_fold_change = ordered_top_genes$logFC

pca_data_projection = scale(datExpr) %*% pca_output$rotation %>% data.frame
p1 = pca_data_projection %>% ggplot(aes(x=PC1, y=PC2, color=ASD_pvals)) + geom_point(alpha=0.5) + theme_minimal()
p2 = pca_data_projection %>% ggplot(aes(x=PC1, y=PC2, color=log_fold_change)) + geom_point() + theme_minimal() +
  scale_colour_gradient2()
grid.arrange(p1, p2, ncol=2)

rm(mod, corfit, lmfit, fit, ASD_pvals, log_fold_change, top_genes, ordered_top_genes, datExpr, datProbes, datSeq, p1, p2)
```

### Clustering

```{r}
clusterings = list()
```
#### K-means Clustering
```{r}
set.seed(123)
wss = sapply(1:10, function(k) kmeans(datExpr_redDim, k, iter.max=100, nstart=25,
                                      algorithm='MacQueen')$tot.withinss)
plot(wss, type='b', main='K-Means Clustering')
best_k = 3
abline(v = best_k, col='blue')

datExpr_k_means = kmeans(datExpr_redDim, best_k, iter.max=100, nstart=25)
clusterings[['km']] = datExpr_k_means$cluster
```

#### Hierarchical Clustering

Chose k=6 as best number of clusters.

Clusters seem to be able to separate ASD and control samples pretty well and there are no noticeable patterns regarding sex, age or brain region in any cluster.

Younger ASD samples seem to be more similar to control samples than older ASD samples (pink cluster has most of the youngest samples). The yellow cluster is made of young ASD samples.

```{r}
h_clusts = datExpr_redDim %>% dist %>% hclust
plot(h_clusts, hang = -1, cex = 0.6, labels=FALSE)
best_k = 6
clusterings[['hc']] = cutree(h_clusts, best_k)
```

#### Consensus Clustering

Samples are grouped into two big clusters and two outliers, the first big cluster has three main subclusters, two small subclusters and two outliers, and the second one two main subclustersand five small subclusters.

*Output plots in clustering_genes_03_20 folder
```{r echo=FALSE, message=FALSE}
# cc_output = datExpr_redDim %>% as.matrix %>% t %>% ConsensusClusterPlus(maxK=8, reps=5, seed=123,
#                                                    title='clustering_genes_03_15/cc_l1/', plot='png')
load('clustering_genes_03_20/cc_l1/cc_output.RData')
best_k = 4 # 2 big clusters and 2 outliers
clusterings[['cc_l1']] = cc_output[[best_k]]$consensusClass = cc_output[[best_k]]$consensusClass

clusterings[['cc_l2']] = clusterings[['cc_l1']]
# cc_output_c1 = datExpr_redDim %>% filter(cc_clusters==1) %>% as.matrix %>% t %>%
#   ConsensusClusterPlus(maxK=8, reps=50, seed=123, title='clustering_samples_03_15/cc_l2_1/', plot='png')
load('clustering_genes_03_20/cc_l2_1/cc_output.RData')
best_k = 7
clusterings[['cc_l2']][clusterings[['cc_l1']]==1] = cc_output_c1[[best_k]]$consensusClass %>%
                                                    sapply(function(x) glue('1_', x))

# cc_output_c2 = datExpr_redDim %>% filter(cc_clusters==2) %>% as.matrix %>% t %>% 
#   ConsensusClusterPlus(maxK=8, reps=50, seed=123, title='clustering_samples_03_15/cc_l2_2/', plot='png')
load('clustering_genes_03_20/cc_l2_2/cc_output.RData')
best_k = 2
clusterings[['cc_l2']][clusterings[['cc_l1']]==2] = cc_output_c2[[best_k]]$consensusClass %>% 
                                                    sapply(function(x) glue('2_', x))
```

#### Independent Component Analysis

Following [this paper's](www.journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1002367) guidelines:

0. Run PCA and keep enough components to explain 60% of the variance

1. Run ICA with that same number of nbComp as principal components kept to then filter them

2. Select components with kurtosis > 3

3. Assign genes to clusters with FDR<0.01 using the fdrtool package

```{r echo=TRUE, results='hide', message=FALSE, warning=FALSE}
ICA_output = datExpr_redDim %>% runICA(nbComp=ncol(datExpr_redDim), method='JADE')
signals_w_kurtosis = ICA_output$S %>% data.frame %>% dplyr::select(names(apply(ICA_output$S, 2, kurtosis)>3))
ICA_clusters = apply(signals_w_kurtosis, 2, function(x) fdrtool(x, plot=F)$qval<0.01) %>% data.frame
ICA_clusters = ICA_clusters[colSums(ICA_clusters)>1]

clusterings[['ICA_min']] = rep(NA, nrow(ICA_clusters))
ICA_clusters_names = colnames(ICA_clusters)
for(c in ICA_clusters_names) clusterings[['ICA_min']][ICA_clusters[,c]] = c
```

Leaves most of the observations (~80%) without a cluster (every time it leaves out more genes):
```{r}
ICA_clusters %>% rowSums %>% table

ICA_clusters %>% mutate(cl_sum=rowSums(.)) %>% as.matrix %>% melt %>% ggplot(aes(Var2,Var1)) + 
  geom_tile(aes(fill=value)) + xlab('Clusters') + ylab('Samples') + 
  theme_minimal() + theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) + coord_flip()
```

#### WGCNA

best_power=56 but blockwiseModules only accepts powers up to 30, so 30 was used instead

```{r, warning=FALSE}
best_power = datExpr_redDim %>% t %>% pickSoftThreshold(powerVector = seq(20, 60, by=2))
network = datExpr_redDim %>% t %>% blockwiseModules(power=30, numericLabels=TRUE)
clusterings[['WGCNA']] = network$colors
names(clusterings[['WGCNA']]) = rownames(datExpr_redDim)
```

It leaves 304 genes without a cluster:
```{r}
clusterings[['WGCNA']] %>% table
```

#### Gaussian Mixture Models with hard thresholding

Number of clusters that resemble more Gaussian mixtures = 39

```{r}
n_clust = datExpr_redDim %>% Optimal_Clusters_GMM(max_clusters=50, criterion='BIC', plot_data=FALSE)
plot(n_clust, type='l', main='Bayesian Information Criterion to choose number of clusters')
best_k = 39
gmm = datExpr_redDim %>% GMM(best_k)
clusterings[['GMM']] = gmm$Log_likelihood %>% apply(1, which.max)
```

Plot of clusters with their centroids in gray
```{r}
gmm_points = rbind(datExpr_redDim, setNames(data.frame(gmm$centroids), names(datExpr_redDim)))
gmm_labels = c(clusterings[['GMM']], rep(NA, best_k)) %>% as.factor
ggplotly(gmm_points %>% ggplot(aes(x=PC1, y=PC2, color=gmm_labels)) + geom_point() + theme_minimal())
```

Trying with 27 clusters (27 has the 9th lowest value)
```{r}
best_k = 27
gmm = datExpr_redDim %>% GMM(best_k)
clusterings[['GMM_2']] = gmm$Log_likelihood %>% apply(1, which.max)
clusterings[['GMM_2']] %>% table
```

Trying with 6 clusters
```{r}
best_k = 6
gmm = datExpr_redDim %>% GMM(best_k)
clusterings[['GMM_3']] = gmm$Log_likelihood %>% apply(1, which.max)
clusterings[['GMM_3']] %>% table
```

#### Manual clustering

Separate the two clouds of points by a straight line. There seems to be a difference in the mean expression of the genes between clusters but not in their standard deviation.
```{r}
manual_clusters = as.factor(as.numeric(-0.1*datExpr_redDim$PC1 + 0.05 > datExpr_redDim$PC2))
datExpr_redDim %>% ggplot(aes(x=PC1, y=PC2, color=manual_clusters)) + geom_point() + 
  geom_abline(slope=-0.1, intercept=0.05, color='gray') + theme_minimal()
names(manual_clusters) = rownames(datExpr_redDim)

clusterings[['Manual']] = manual_clusters
```

Cluster 2 has a bigger mean expression than cluster 1. There also seem to be more than one distributions in cluster 2, with two different means and three different standard deviations?
```{r, fig.width = 12}
manual_clusters_data = cbind(apply(datExpr_redDim, 1, mean), apply(datExpr_redDim, 1, sd), 
                             manual_clusters) %>% data.frame
colnames(manual_clusters_data) = c('mean','sd','cluster')
manual_clusters_data = manual_clusters_data %>% mutate('cluster'=as.factor(cluster))
p1 = manual_clusters_data %>% ggplot(aes(x=mean, color=cluster, fill=cluster)) + 
  geom_density(alpha=0.4) + theme_minimal()
p2 = manual_clusters_data %>% ggplot(aes(x=sd, color=cluster, fill=cluster)) + 
  geom_density(alpha=0.4) + theme_minimal()
grid.arrange(p1, p2, ncol=2)
```

Separating cluster 2 genes by their sd:
```{r}
c2_sd = manual_clusters_data %>% filter(cluster==2) %>% dplyr::select(sd)
rownames(c2_sd) = rownames(manual_clusters_data)[manual_clusters_data$cluster=='2']
gmm_c2_sd = c2_sd %>% GMM(3)

manual_clusters_data %>% ggplot(aes(x=sd)) +
  stat_function(fun=dnorm, n=100, colour='#99cc00', # green
                args=list(mean=gmm_c2_sd$centroids[1], sd=gmm_c2_sd$covariance_matrices[1])) +
  stat_function(fun=dnorm, n=100, colour='#2eb8b8', # acqua
                args=list(mean=gmm_c2_sd$centroids[2], sd=gmm_c2_sd$covariance_matrices[2])) +
  stat_function(fun=dnorm, n=100, colour='#a64dff', # purple
                args=list(mean=gmm_c2_sd$centroids[3], sd=gmm_c2_sd$covariance_matrices[3])) +
  theme_minimal()

clusterings[['Manual2']] = as.character(clusterings[['Manual']])
clusterings[['Manual2']][clusterings[['Manual']]==1] = gmm_c2_sd$Log_likelihood %>% 
  apply(1, function(x) glue('1_',which.max(x)))

clusterings[['Manual2']] %>% table
```

```{r, warning=FALSE}
rm(wss, datExpr_k_means, h_clusts, cc_output, cc_output_c1, cc_output_c2, best_k, ICA_output, 
   ICA_clusters_names, signals_w_kurtosis, n_clust, gmm, gmm_points, gmm_labels, network, 
   best_power, c, manual_clusters, manual_clusters_data, c2_sd, gmm_c2_sd, p1, p2, 
   pca_data_projection)
```

### Compare clusterings

Using Adjusted Rand Index:

* Clusterings seem to give very different results and none resembles the manual separation

* Simple methods give similar results (K-means, hierarchical clustering, consensus clustering)
```{r}
cluster_sim = data.frame(matrix(nrow = length(clusterings), ncol = length(clusterings)))
for(i in 1:(length(clusterings))){
  cluster1 = as.factor(clusterings[[i]])
  for(j in (i):length(clusterings)){
    cluster2 = as.factor(clusterings[[j]])
    cluster_sim[i,j] = adj.rand.index(cluster1, cluster2)
  }
}
colnames(cluster_sim) = names(clusterings)
rownames(cluster_sim) = colnames(cluster_sim)

cluster_sim = cluster_sim %>% as.matrix %>% round(2)
heatmap.2(x = cluster_sim, Rowv = FALSE, Colv = FALSE, dendrogram = 'none', 
          cellnote = cluster_sim, notecol = 'black', trace = 'none', key = FALSE, 
          cexRow = 1, cexCol = 1, margins = c(7,7))
 

rm(i, j, cluster1, cluster2, cluster_sim)
```

### Scatter plots

```{r}
create_2D_plot = function(cat_var){
 ggplotly(plot_points %>% ggplot(aes_string(x='PC1', y='PC2', color=cat_var)) + 
          geom_point() + theme_minimal() + 
          xlab(paste0('PC1 (', round(summary(pca_output)$importance[2,1]*100,2),'%)')) +
          ylab(paste0('PC2 (', round(summary(pca_output)$importance[2,2]*100,2),'%)')))
}
create_3D_plot = function(cat_var){
  plot_points %>% plot_ly(x=~PC1, y=~PC2, z=~PC3) %>% add_markers(color=plot_points[,cat_var], size=1) %>% 
    layout(title = glue('Samples coloured by ', cat_var),
           scene = list(xaxis=list(title=glue('PC1 (',round(summary(pca_output)$importance[2,1]*100,2),'%)')),
                        yaxis=list(title=glue('PC2 (',round(summary(pca_output)$importance[2,2]*100,2),'%)')),
                        zaxis=list(title=glue('PC3 (',round(summary(pca_output)$importance[2,3]*100,2),'%)'))))  
}

plot_points = datExpr_redDim %>% data.frame() %>% dplyr::select(PC1:PC3) %>%
              mutate(ID = rownames(.),                           km_clust = as.factor(clusterings[['km']]),
                hc_clust = as.factor(clusterings[['hc']]),       cc_l1_clust = as.factor(clusterings[['cc_l1']]),
                cc_clust = as.factor(clusterings[['cc_l2']]),    ica_clust = as.factor(clusterings[['ICA_min']]),
                n_ica_clust = as.factor(rowSums(ICA_clusters)),  gmm_clust = as.factor(clusterings[['GMM']]),
                gmm_clust2 = as.factor(clusterings[['GMM_2']]),  gmm_clust3 = as.factor(clusterings[['GMM_3']]),
                wgcna_clust = as.factor(clusterings[['WGCNA']]), manual_clust=as.factor(clusterings[['Manual']]),
                manual2_clust=as.factor(clusterings[['Manual2']]))
```

#### 2D plots of clusterings

* Simple methods seem to only partition the space in buckets using information from the first component

* All clusterings except K-means manage to separate the two clouds at least partially, but no one does a good job

* WGCNA creates clusters inverted between clouds

* Manual classification + GMM by standard deviation clusters make sense
```{r, warning=FALSE}
create_2D_plot('km_clust')
create_2D_plot('hc_clust')
create_2D_plot('cc_l1_clust')
create_2D_plot('cc_clust')
create_2D_plot('ica_clust')
create_2D_plot('gmm_clust')
create_2D_plot('gmm_clust2')
create_2D_plot('gmm_clust3')
create_2D_plot('wgcna_clust')
create_2D_plot('manual2_clust')
```

#### 3D plots
```{r, warning=FALSE}
create_3D_plot('ica_clust')
create_3D_plot('gmm_clust')
create_3D_plot('gmm_clust3')
create_3D_plot('wgcna_clust')
create_3D_plot('manual2_clust')
```