---
title: "Clustering Genes"
# author: "Magdalena Navarro"
date: "15 March 2019"
output:
  html_document:
    code_folding: "hide"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Filter criteria:

- Filter differentially expressed genes between autism and control (p-value < 0.05)

- No samples are removed based on network connectivity z-scores

```{r, echo=TRUE, include=FALSE}
setwd('/home/magdalena/PhD/initialExperiments/Gandal/RNAseq/R_markdowns')

library(limma);
library(tidyverse) ; library(reshape2) ; library(glue) ; library(plotly) ; library(dendextend)
library(RColorBrewer) ; library(viridis)
library(ConsensusClusterPlus)
library(JADE) ; library(MineICA) ; library(moments) ; library(fdrtool)
library(ClusterR)
library(WGCNA)
library(pdfCluster) ; library(gplots)

# # FILTER DE GENES
# load('./../working_data/RNAseq_ASD_4region_normalized.Rdata')
# 
# # Balance Groups by covariates, remove singular batches (none)
# to_keep = (datMeta$Subject_ID != 'AN03345') & !is.na(datMeta$Dx)
# table(to_keep)
# datMeta = datMeta[to_keep,]
# datExpr = datExpr[,to_keep]
# datSeq = datSeq[to_keep,]
# 
# # Select genes differentially expressed in ASD
# mod = model.matrix(~ Dx, data=datMeta)
# corfit = duplicateCorrelation(datExpr, mod, block=datMeta$Subject_ID)
# lmfit = lmFit(datExpr, design=mod, block=datMeta$Subject_ID, correlation=corfit$consensus)
# fit = eBayes(lmfit, trend=T, robust=T)
# ASD_pvals = fit$p.value[,'DxASD']
# 
# to_keep = ASD_pvals<0.05
# table(to_keep)/length(to_keep)*100 # keep 32% of genes
# datExpr = datExpr[to_keep,]
#
# rm(corfit, fit, lmfit, mod, ASD_pvals, to_keep)

load('./../working_data/RNAseq_ASD_4region_DEgenes.Rdata')
```

```{r}
glue('Number of genes: ', nrow(datExpr), '\n',
     'Number of samples: ', ncol(datExpr), ' (', sum(datMeta$Diagnosis_=='ASD'), ' ASD, ',
     sum(datMeta$Diagnosis_!='ASD'), ' controls)')
```

### Dimensionality reduction using PCA

First principal component explains over 90% of the total variance

```{r}
reduce_dim_datExpr = function(datExpr, datMeta, var_explained=0.98){
  
  datExpr = data.frame(datExpr)
  
  datExpr_pca = prcomp(datExpr, scale.=TRUE)
  par(mfrow=c(1,2))
  plot(summary(datExpr_pca)$importance[2,], type='b')
  plot(summary(datExpr_pca)$importance[3,], type='b')
  abline(h=var_explained, col='blue')
  
  last_pc = data.frame(summary(datExpr_pca)$importance[3,]) %>% rownames_to_column(var='ID') %>% 
    filter(.[[2]] >= var_explained) %>% top_n(-1, ID)
  
  print(glue('Keeping top ', substr(last_pc$ID, 3, nchar(last_pc$ID)), ' components that explain ',
             var_explained*100, '% of the variance'))
  
  datExpr_top_pc = datExpr_pca$x %>% data.frame %>% dplyr::select(PC1:last_pc$ID)
  
  return(list('datExpr'=datExpr_top_pc, 'pca_output'=datExpr_pca))
}

reduce_dim_output = reduce_dim_datExpr(datExpr, datMeta)
datExpr_redDim = reduce_dim_output$datExpr
pca_output = reduce_dim_output$pca_output

rm(datSeq, datProbes, reduce_dim_datExpr, reduce_dim_output, datExpr)
```

Genes seem to have separated into two clouds of points:
```{r}
datExpr_redDim %>% ggplot(aes(x=PC1, y=PC2)) + geom_point() + theme_minimal()
```

Projecting all the original points into the space created by the two principal components and colouring by the differential expression p-value we can see that the points in the middle of the two clouds were filtered out because their DE wasn't statistically significant
```{r}
load('./../working_data/RNAseq_ASD_4region_normalized.Rdata')

# Remove Subject with ID = 'AN03345'
keep = datMeta$Subject_ID!='AN03345'
datMeta = datMeta[keep,]
datExpr = datExpr[,keep]

# Calculate differential expression p-value of each gene
mod = model.matrix(~ Dx, data=datMeta)
corfit = duplicateCorrelation(datExpr, mod, block=datMeta$Subject_ID)
lmfit = lmFit(datExpr, design=mod, block=datMeta$Subject_ID, correlation=corfit$consensus)
fit = eBayes(lmfit, trend=T, robust=T)
ASD_pvals = fit$p.value[,'DxASD']

pca_data_projection = scale(datExpr) %*% pca_output$rotation %>% data.frame
pca_data_projection %>% ggplot(aes(x=PC1, y=PC2, color=ASD_pvals)) + geom_point(alpha=0.5) + theme_minimal()

rm(mod, corfit, lmfit, fit, ASD_pvals, datExpr, datProbes, datSeq)
```

### Clustering

```{r}
clusterings = list()
```
#### K-means Clustering
```{r}
set.seed(123)
wss = sapply(1:10, function(k) kmeans(datExpr_redDim, k, iter.max=100, nstart=25,
                                      algorithm='MacQueen')$tot.withinss)
plot(wss, type='b', main='K-Means Clustering')
best_k = 4
abline(v = best_k, col='blue')

datExpr_k_means = kmeans(datExpr_redDim, best_k, nstart=25)
clusterings[['km']] = datExpr_k_means$cluster
```

#### Hierarchical Clustering

Chose k=6 as best number of clusters.

Clusters seem to be able to separate ASD and control samples pretty well and there are no noticeable patterns regarding sex, age or brain region in any cluster.

Younger ASD samples seem to be more similar to control samples than older ASD samples (pink cluster has most of the youngest samples). The yellow cluster is made of young ASD samples.


```{r}
h_clusts = datExpr_redDim %>% dist %>% hclust
plot(h_clusts, hang = -1, cex = 0.6, labels=FALSE)
best_k = 5
clusterings[['hc']] = cutree(h_clusts, best_k)
```

#### Consensus Clustering

Samples are grouped into two big clusters and then each cluster in 4 and 6 subclusters, respectively. The first big separation into two clusters is very clear, the subclusters not so much.

*Output plots in clustering_genes_03_15 folder
```{r echo=FALSE, message=FALSE}
# cc_output = datExpr_redDim %>% as.matrix %>% t %>% ConsensusClusterPlus(maxK=8, reps=5, seed=123,
#                                                    title='clustering_genes_03_15/cc_l1/', plot='png')
load('clustering_genes_03_15/cc_l1/cc_output.RData')
best_k = 5 # 2 clusters and 3 outliers
clusterings[['cc_l1']] = cc_output[[best_k]]$consensusClass = cc_output[[best_k]]$consensusClass

clusterings[['cc_l2']] = clusterings[['cc_l1']]
# cc_output_c1 = datExpr_redDim %>% filter(cc_clusters==1) %>% as.matrix %>% t %>% 
#   ConsensusClusterPlus(maxK=8, reps=50, seed=123, title='clustering_samples_03_15/cc_l2_1/', plot='png')
load('clustering_genes_03_15/cc_l2_1/cc_output.RData')
best_k = 4
clusterings[['cc_l2']][clusterings[['cc_l2']]==1] = cc_output_c1[[best_k]]$consensusClass %>% 
                                                    sapply(function(x) glue('1_', x))

# cc_output_c2 = datExpr_redDim %>% filter(cc_clusters==2) %>% as.matrix %>% t %>% 
#   ConsensusClusterPlus(maxK=8, reps=50, seed=123, title='clustering_samples_03_15/cc_l2_2/', plot='png')
load('clustering_genes_03_15/cc_l2_2/cc_output.RData')
best_k = 6
clusterings[['cc_l2']][clusterings[['cc_l2']]==2] = cc_output_c2[[best_k]]$consensusClass %>% 
                                                    sapply(function(x) glue('2_', x))
```

#### Independent Component Analysis

Following [this paper's](www.journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1002367) guidelines:

0. Run PCA and keep enough components to explain 60% of the variance

1. Run ICA with that same number of nbComp as principal components kept to then filter them

2. Select components with kurtosis > 3

3. Assign genes to clusters with FDR<0.001 using the fdrtool package

```{r echo=TRUE, results='hide', message=FALSE, warning=FALSE}
ICA_output = datExpr_redDim %>% runICA(nbComp=ncol(datExpr_redDim), method='JADE')
signals_w_kurtosis = ICA_output$S %>% data.frame %>% dplyr::select(names(apply(ICA_output$S, 2, kurtosis)>3))
ICA_clusters = apply(signals_w_kurtosis, 2, function(x) fdrtool(x, plot=F)$qval<0.01) %>% data.frame
ICA_clusters = ICA_clusters[colSums(ICA_clusters)>1]

clusterings[['ICA_min']] = rep(NA, nrow(ICA_clusters))
ICA_clusters_names = colnames(ICA_clusters)
for(c in ICA_clusters_names) clusterings[['ICA_min']][ICA_clusters[,c]] = c
```

Leaves most of the observations (~75%) without a cluster:
```{r}
ICA_clusters %>% rowSums %>% table

ICA_clusters %>% mutate(cl_sum=rowSums(.)) %>% as.matrix %>% melt %>% ggplot(aes(Var2,Var1)) + 
  geom_tile(aes(fill=value)) + xlab('Clusters') + ylab('Samples') + 
  theme_minimal() + theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) + coord_flip()
```

#### WGCNA

best_power=40 but blockwiseModules only accepts powers up to 30, so 30 was used instead

```{r, warning=FALSE}
best_power = datExpr_redDim %>% t %>% pickSoftThreshold(powerVector = seq(30, 50, by=1))
network = datExpr_redDim %>% t %>% blockwiseModules(power=30, numericLabels=TRUE)
clusterings[['WGCNA']] = network$colors
names(clusterings[['WGCNA']]) = rownames(datExpr_redDim)
```

It leaves 476 genes without a cluster:
```{r}
table(clusterings[['WGCNA']])
```

#### Gaussian Mixture Models with hard thresholding

Number of clusters that resemble more Gaussian mixtures = 32 but perhaps a smaller number could also work (maybe 5, 12 or 17)

```{r}
n_clust = datExpr_redDim %>% Optimal_Clusters_GMM(max_clusters=50, criterion='BIC', plot_data=FALSE)
plot(n_clust, type='l', main='Bayesian Information Criterion to choose number of clusters')
best_k = 32
gmm = datExpr_redDim %>% GMM(best_k)
clusterings[['GMM']] = gmm$Log_likelihood %>% apply(1, which.max)
```

Plot of clusters with their centroids in gray
```{r}
gmm_points = rbind(datExpr_redDim, setNames(data.frame(gmm$centroids), names(datExpr_redDim)))
gmm_labels = c(clusterings[['GMM']], rep(NA, best_k)) %>% as.factor
ggplotly(gmm_points %>% ggplot(aes(x=PC1, y=PC2, color=gmm_labels)) + geom_point() + theme_minimal())
```

#### Manual clustering

Separate the two clouds of points by a straight line. There seems to be a difference in the mean expression of the genes between clusters but not in their standard deviation.
```{r}
manual_clusters = as.factor(as.numeric(0.08*datExpr_redDim$PC1 + 0.2 > datExpr_redDim$PC2))
datExpr_redDim %>% ggplot(aes(x=PC1, y=PC2, color=manual_clusters)) + geom_point() + 
  geom_abline(slope=0.08, intercept=0.2, color='gray') + theme_minimal()
names(manual_clusters) = rownames(datExpr_redDim)

clusterings[['Manual']] = manual_clusters

manual_clusters_data = cbind(apply(datExpr_redDim, 1, mean), apply(datExpr_redDim, 1, sd), 
                             manual_clusters) %>% data.frame
colnames(manual_clusters_data) = c('mean','sd','cluster')
manual_clusters_data = manual_clusters_data %>% mutate('cluster'=as.factor(cluster))
manual_clusters_data %>% ggplot(aes(x=mean, color=cluster, fill=cluster)) + 
  geom_density(alpha=0.4) + theme_minimal()
manual_clusters_data %>% ggplot(aes(x=sd, color=cluster, fill=cluster)) + 
  geom_density(alpha=0.4) + theme_minimal()
```

```{r}
rm(wss, datExpr_k_means, h_clusts, cc_output, cc_output_c1, cc_output_c2, best_k, ICA_output, 
   ICA_clusters_names, signals_w_kurtosis, n_clust, gmm, gmm_points, gmm_labels, network, 
   best_power, c, manual_clusters, manual_clusters_data)
```

### Compare clusterings

Using Adjusted Rand Index:

* Clusterings seem to give very different results and none resembles the manual separation

* Simple methods give similar results (K-means, hierarchical clustering, consensus clustering)
```{r}
cluster_sim = data.frame(matrix(nrow = length(clusterings), ncol = length(clusterings)))
for(i in 1:(length(clusterings))){
  cluster1 = as.factor(clusterings[[i]])
  for(j in (i):length(clusterings)){
    cluster2 = as.factor(clusterings[[j]])
    cluster_sim[i,j] = adj.rand.index(cluster1, cluster2)
  }
}
colnames(cluster_sim) = names(clusterings)
rownames(cluster_sim) = colnames(cluster_sim)

cluster_sim = cluster_sim %>% as.matrix %>% round(2)
heatmap.2(x = cluster_sim, Rowv = FALSE, Colv = FALSE, dendrogram = 'none', 
          cellnote = cluster_sim, notecol = 'black', trace = 'none', key = FALSE, 
          cexRow = 1, cexCol = 1, margins = c(7,7))
 

rm(i, j, cluster1, cluster2, cluster_sim)
```

### Scatter plots

```{r}
create_2D_plot = function(cat_var){
 ggplotly(plot_points %>% ggplot(aes_string(x='PC1', y='PC2', color=cat_var)) + 
          geom_point() + theme_minimal() + 
          xlab(paste0('PC1 (', round(summary(pca_output)$importance[2,1]*100,2),'%)')) +
          ylab(paste0('PC2 (', round(summary(pca_output)$importance[2,2]*100,2),'%)')))
}
create_3D_plot = function(cat_var){
  plot_points %>% plot_ly(x=~PC1, y=~PC2, z=~PC3) %>% add_markers(color=plot_points[,cat_var], size=1) %>% 
    layout(title = glue('Samples coloured by ', cat_var),
           scene = list(xaxis=list(title=glue('PC1 (',round(summary(pca_output)$importance[2,1]*100,2),'%)')),
                        yaxis=list(title=glue('PC2 (',round(summary(pca_output)$importance[2,2]*100,2),'%)')),
                        zaxis=list(title=glue('PC3 (',round(summary(pca_output)$importance[2,3]*100,2),'%)'))))  
}

plot_points = datExpr_redDim %>% data.frame() %>% dplyr::select(PC1:PC3) %>%
              mutate(ID = rownames(.),                           km_clust = as.factor(clusterings[['km']]),
                hc_clust = as.factor(clusterings[['hc']]),       cc_l1_clust = as.factor(clusterings[['cc_l1']]),
                cc_clust = as.factor(clusterings[['cc_l2']]),    ica_clust = as.factor(clusterings[['ICA_min']]),
                n_ica_clust = as.factor(rowSums(ICA_clusters)),  gmm_clust = as.factor(clusterings[['GMM']]),
                wgcna_clust = as.factor(clusterings[['WGCNA']]), manual_clust=as.factor(clusterings[['Manual']]))
```

2D plots of clusterings

* Simple methods seem to only partition the space in buckets using information from the first component

* ICA seems to be the only one to distinguish between the two clouds of points, although it is really noisy

* WGCNA creates clusters inverted between clouds
```{r}
create_2D_plot('km_clust')
create_2D_plot('hc_clust')
create_2D_plot('cc_l1_clust')
create_2D_plot('cc_clust')
create_2D_plot('ica_clust')
create_2D_plot('gmm_clust')
create_2D_plot('wgcna_clust')
```

3D plots
```{r}
create_3D_plot('ica_clust')
```